### 2010-04-15 doc/paper.tex
added a basic .tex skeleton
git-svn-id: http://svn.scesoc.ca/evolution/trunk@64 fee98b4c-d1c3-4d34-9525-a468159e4a48

% ===========================================================================
% Title:
% ---------------------------------------------------------------------------
% to create Type I fonts type "dvips -P cmz -t letter <filename>"
% ===========================================================================
\documentclass[11pt]{article}       %--- LATEX 2e base
\usepackage{latexsym}               %--- LATEX 2e base
%---------------- Wide format -----------------------------------------------
\textwidth=6in \textheight=9in \oddsidemargin=0.25in
\evensidemargin=0.25in \topmargin=-0.5in
%--------------- Algorithm --------------------------------------------------
\newtheorem{algX}{Algorithm}
\newenvironment{algorithm}       {\begin{algX}\begin{em}}%
                                 {\par\noindent --- End of Algorithm ---
                                 \end{em}\end{algX}}
\newcommand{\step}[2]            {\begin{list}{}
                                  {  \setlength{\topsep}{0cm}
                                     \setlength{\partopsep}{0cm}
                                     \setlength{\leftmargin}{0.8cm}
                                     \setlength{\labelwidth}{0.7cm}
                                     \setlength{\labelsep}{0.1cm}    }
                                  \item[#1]#2    \end{list}}
                                 % usage: \begin{algorithm} \label{xyz}
                                 %        ... \step{(1)}{...} ...
                                 %        \end{algorithm}
%--------------- Figures ----------------------------------------------------
\usepackage{graphicx}

\newcommand{\includeFig}[3]      {\begin{figure}[htb] \begin{center}
                                 \includegraphics
                                 [width=4in,keepaspectratio]
                                 {#2}\caption{\label{#1}#3} \end{center} \end{figure}}
                                 % usage: \includeFig{label}{file}{caption}

% ===========================================================================
\begin{document}
% ===========================================================================

% ############################################################################
% Title
% ############################################################################

\title{Accelerating Genetic ANN Training using the CUDA Platform}

% ############################################################################
% Author(s) (no blank lines !)
\author{
% ############################################################################
Catalin Patulea and Robert Peace\\
School of Systems and Computer Engineering\\
Carleton University\\
Ottawa, Canada K1S 5B6\\
{\em rpeace@sce.carleton.ca}
% ############################################################################
} % end-authors
% ############################################################################

\maketitle

% ############################################################################
% Abstract
% ############################################################################
\begin{abstract}
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this
\end{abstract}

% ############################################################################
\section{Introduction} \label{intro}
% ############################################################################
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this

% ############################################################################
\section{Background} \label{background}
% ############################################################################
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this

% ----------------------------------------------------------------------------
\subsection{Artificial Neural Networks} \label{ann}
% ----------------------------------------------------------------------------
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this

% ----------------------------------------------------------------------------
\subsection{Genetic Algorithms} \label{ga}
% ----------------------------------------------------------------------------
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this

% ----------------------------------------------------------------------------
\subsection{The CUDA Platform} \label{cuda}
% ----------------------------------------------------------------------------
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this

% ############################################################################
\section{Methods} \label{algimp}
% ############################################################################
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this

% ----------------------------------------------------------------------------
\subsection{Preliminary Data Analysis} \label{prelim}
% ----------------------------------------------------------------------------
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this

% ----------------------------------------------------------------------------
\subsection{Data Preprocessing Steps} \label{preprocessing}
% ----------------------------------------------------------------------------
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this

% ----------------------------------------------------------------------------
\subsection{Classifier Design} \label{design}
% ----------------------------------------------------------------------------
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this

% ----------------------------------------------------------------------------
\subsection{Classifier Implementation} \label{implementation}
% ----------------------------------------------------------------------------
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this

% ############################################################################
\section{Results} \label{results}
% ############################################################################
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this

% ----------------------------------------------------------------------------
\subsection{Experimental Setup} \label{experiment}
% ----------------------------------------------------------------------------
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this

% ----------------------------------------------------------------------------
\subsection{Performance Results} \label{performance}
% ----------------------------------------------------------------------------
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this

% ----------------------------------------------------------------------------
\subsection{Classifier Results} \label{results}
% ----------------------------------------------------------------------------
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this

% ############################################################################
\section{Conclusion} \label{concl}
% ############################################################################
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this

% ############################################################################
\section{Discussion} \label{disc}
% ############################################################################
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this

% ----------------------------------------------------------------------------
\subsection{Future Work} \label{future}
% ----------------------------------------------------------------------------
Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this

% ############################################################################
% Bibliography
% ############################################################################
\bibliographystyle{plain}
\bibliography{bibliography}     %loads bibliography.bib

% ============================================================================
\end{document}
% ============================================================================

### 2010-04-15 doc/paper.tex
Add Introduction
git-svn-id: http://svn.scesoc.ca/evolution/trunk@65 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 % ############################################################################
-Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this
+Artificial neural networks (ANNs) are a class of very flexible machine learning models applicable to many classification and regression tasks.
+All artificial neural networks share the same basic structure but their adaptability lies in the ability to vary the network topology, activation functions at each node, and the weights between each nodes.
+Together, these factors allow for a very comprehensive tool capable of modeling a wide range of systems.
+
+Generally, for any given problem, the ANN designers choose a particular network structure.
+Then, during supervised learning, the labelled training instances are used to adjust the network parameters (edge weights and other node parameters) until some a given criteria is reached regarding classification error or training time.
+This can be seen as a search for a global minimum in the space of network parameters.
+Because the number of parameters in a typical neural network is in the tens or hundreds, the search is in a quite high-dimensional space and thus quite challenging.
+
+The ANN parameter finding problem can be viewed as an optimization problem.
+Simple algorithms like the first order gradient descent, second-order Gauss-Newton or hybrid Levenberg–Marquardt rely on approximations of the error function to iteratively find the global extremum.
+These algorithms perform well when the error landscape is relatively simple, but for ANNs with complex structures and landscapes with many local extrema, they are susceptible to getting stuck at those local extrema and sometimes do not converge nicely.
+Additionally, they impose certain restrictions on the node function (differentiability).
+
+Stochastic approaches to optimization provide some hope of escaping these shortcomings.
+One such example, simulated annealing, begins with a particular set of parameters and randomly perturbs them in the hopes of approaching an extremum.
+The new solution is chosen probabilistically, depending on a temperature parameter which decreases during this process.
+Thus, early during optimization we favor large-scale exploration of the search space while near the end we are exploring only the neighborhood of a particular local extremum (which we hope is the global extremum) to improve the precision of our solution.
+
+Genetic algorithms (GAs) are stochastic optimization algorithms which try to leverage concepts of evolution and natural selection to find the optimal solution.
+Similarly to simulated annealing, GAs use random perturbations to explore the search space.
+However, GAs additioally operate on multiple solutions at a time and attempt to combine multiple partial solutions into a new better performing candidate.
+At the core of GAs is the definition of a "chromosome", which represents one candidate solution, as well as an associated set of genetic operators which perform domain-specific mutations on this candidate and enable candidate mating.
+
+TODO: talk about GA resource-intensive (esp with ANN)
+TODO: talk about CUDA being good for number-crunching
+TODO: talk about CUDA performance dependent on memory access patterns
 

### 2010-04-16 ann.py
Implement normalization in the Input class
git-svn-id: http://svn.scesoc.ca/evolution/trunk@66 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@ -12,8 +12,58 @@ import math
 
 log = logging.getLogger("ann")
 
+class Parameters(ctypes.Structure):
+  _fields_ = [("ih", 4 * (19 * ctypes.c_float)),
+              ("c", 4 * (19 * ctypes.c_float)),
+              ("w", 4 * ctypes.c_float),
+              ("ho", 4 * ctypes.c_float)]
+
+  def _float_list_str(self, l):
+    return ", ".join("%4.01g" % el for el in l)
+
+  def __str__(self):
+    s = []
+    s.append("Parameters(\n")
+    for i in range(4):
+      ihl = list(self.ih[i])
+      s.append("  ih[%d]=[0: %s,\n" % (i, self._float_list_str(ihl[0:10])))
+      s.append("        10: %s]\n" % (self._float_list_str(ihl[10:])))
+    for i in range(4):
+      cl = list(self.c[i])
+      s.append("  c[%d] =[0: %s,\n" % (i, self._float_list_str(cl[0:10])))
+      s.append("        10: %s]\n" % (self._float_list_str(cl[10:])))
+    s.append("  w =%s,\n" % self._float_list_str(list(self.w)))
+    s.append("  ho=%s,\n" % self._float_list_str(list(self.ho)))
+    s.append(")")
+    return "".join(s)
+
+  def _array_repr(self, dims, value):
+    if dims == []:
+      return "%e" % value
+
+    typestr = "ctypes.c_float"
+    for dim in reversed(dims):
+      typestr = "(%d*%s)" % (dim, typestr)
+    
+    valuestr = typestr + "(" + ", ".join(self._array_repr(dims[1:], subvalue) for subvalue in value) + ")"
+    return valuestr
+
+  def __repr__(self):
+    r = ["Parameters(\n"]
+    r.append("  ih=%s,\n" % self._array_repr([4, 19], self.ih))
+    r.append("  c=%s,\n" % self._array_repr([4, 19], self.c))
+    r.append("  w=%s,\n" % self._array_repr([4], self.w))
+    r.append("  ho=%s\n" % self._array_repr([4], self.ho))
+    r.append(")")
+    return "".join(r)
+
+  @staticmethod
+  def from_file(annfile):
+    return eval(compile(open(annfile).read(), annfile, "eval"),
+                {"Parameters": Parameters,
+                 "ctypes": ctypes})
+
[…]

### 2010-04-16 ann_kernels.cu

@@ -2,13 +2,12 @@
 #include <math_constants.h>
 
 #define ARRAY_SIZE(array) (sizeof(array) / sizeof((array)[0]))
-#define NODES_PER_LAYER 6
 
 struct Parameters {
-  float ih[NODES_PER_LAYER][19];  // input->hidden edge weight
-  float c[NODES_PER_LAYER][19];   // RBF center
-  float w[NODES_PER_LAYER];       // RBF width
-  float ho[NODES_PER_LAYER];      // hidden->output edge weight
+  float ih[4][19];  // input->hidden edge weight
+  float c[4][19];   // RBF center
+  float w[4];       // RBF width
+  float ho[4];      // hidden->output edge weight
 };
 
 __global__ void evaluate(
@@ -36,7 +35,7 @@ __global__ void evaluate(
     float output = 0.f;
     
     // TODO: Make sure this gets unrolled
-    for (int j = 0; j < NODES_PER_LAYER; j++) {
+    for (int j = 0; j < 4; j++) {
       float d2 = 0.f; // input to hidden node j
       for (int i = 0; i < 19; i++) {
         const float d = inputs[i] * p->ih[j][i] - p->c[j][i];

### 2010-04-16 blind.py

@@ -21,9 +21,10 @@ def main(annfile):
   for rank0, (index, output) in enumerate(sortedOutputs):
     indexToRank[index] = rank0 + 1
 
-  for index, output in enumerate(outputs):
+  originalBlindSet = open("test.tsv")
+  for index, (row, output) in enumerate(zip(originalBlindSet, outputs)):
     rank = indexToRank[index]
-    print index, output, rank
+    print "%s\t%d" % (row.strip(), rank)
 
 if __name__ == "__main__":
   if len(sys.argv) != 2:

### 2010-04-16 classify.py

@@ -1,5 +1,5 @@
 #!/usr/bin/python
-from ann2layer import Parameters, ANN
+from ann import Parameters, ANN
 import input
 import sys
 import ctypes
@@ -63,7 +63,7 @@ class SampleTester(object):
 def main(annfile):
   randSample = random.Random(input.SAMPLE_SEED)
   
-  inp = input.Input("train3-std.tsv", randSample)
+  inp = input.Input("train3.tsv", randSample)
   
   print "Test set:",
   inp.testSet.show()

### 2010-04-16 genops.py

@@ -21,7 +21,7 @@ Population generation / initializer.  Uses method from Montana and Davis.
 def generatePop(generation):
     for i in range(POPSIZE):
         nextMember = Parameters()
-        for j in range(ANN.NODES_PER_LAYER):
+        for j in range(4):
             for k in range(19):
                 nextMember.ih[j][k] = getInitialFloat()
                 nextMember.c[j][k] = getInitialFloat()
@@ -39,14 +39,14 @@ Create a new generation, based on the current generation
 @type fitList: A list of float values
 @return: List of new parameters (IE, a new generation)
 """
-def generateGeneration(oldGen, mutateValue):
+def generateGeneration(oldGen):
     newGen = []
     newGen.append(oldGen[0])
     for i in range(POPSIZE - 1):
-        if 10*random.random() > mutateValue + 2.5:
+        if random.choice([0,1]) == 1:
             index = getIndex()
             newGen.append(
-                mutate(oldGen[index], mutateValue)
+                mutate(oldGen[index])
             )
         else:
             index = getIndex()
@@ -66,22 +66,22 @@ Mutation operator.  Uses MUTATE NODES operator from Montana and Davis.
 @type xman: Parameters (see class Parameters, above)
 @return Mutated Parameters object
 """
-def mutate(xman, mutateValue):  
+def mutate(xman):  
     xmanJr = Parameters()
 
-    for i in range(ANN.NODES_PER_LAYER):
+    for i in range(4):
         for j in range(19):
             xmanJr.ih[i][j] = xman.ih[i][j]
             xmanJr.c[i][j] = xman.c[i][j]
         xmanJr.w[i] = xman.w[i]
         xmanJr.ho[i] = xman.ho[i]
 
-    node = random.randint(0,ANN.NODES_PER_LAYER-1)
+    node = random.randint(0,3)
     for i in range(19):
-        xmanJr.ih[node][i] += getMutationValue(mutateValue)
-        xmanJr.c[node][i] += getMutationValue(mutateValue)
-    xmanJr.w[node] += getMutationValue(mutateValue)
-    xmanJr.ho[node] += getMutationValue(mutateValue)
+        xmanJr.ih[node][i] += getMutationValue()
+        xmanJr.c[node][i] += getMutationValue()
+    xmanJr.w[node] += getMutationValue()
+    xmanJr.ho[node] += getMutationValue()
     
     return xmanJr
 
[…]

### 2010-04-16 genplot.py

@@ -1,10 +1,51 @@
-import matplotlib.pyplot as plt
+from pygooglechart import XYLineChart, Axis
+import cgi
 import numpy
 
-def plot():
-    plt.show()
+minFitnessHistory = []
+avgFitnessHistory = []
+maxFitnessHistory = []
+testFitnessHistory = []
 
-def addGeneration(fitnessList, testFit, genNumber):
-    plt.scatter(genNumber, max(fitnessList))
-    plt.scatter(genNumber, testFit, marker="x")
-    plt.scatter(genNumber, min(fitnessList))
+def plot():
+    pass
+
+def addGeneration(fitnessList, testFitness, genNumber):
+    fmin = min(fitnessList)
+    avg = sum(fitnessList) / len(fitnessList)
+    fmax = max(fitnessList)
+    print "G%d: avg %.02f min %.02f max %.02f test %.02f" % (
+        genNumber,
+        avg,
+        fmin,
+        fmax,
+        testFitness
+    )
+
+    minFitnessHistory.append(fmin)
+    avgFitnessHistory.append(avg)
+    maxFitnessHistory.append(fmax)
+    testFitnessHistory.append(testFitness)
+    
+    #xMax = len(avgFitnessHistory) * 120/100
+    xMax = 120
+    
+    chart = XYLineChart(640, 400, x_range=(0, xMax), y_range=(0, 6),
+      legend=["min", "avg", "max", "test"],
+      colours=["66CC00", "FF9933", "FF0000", "0000FF"])
+    chart.set_axis_range(Axis.BOTTOM, 0, xMax)
+    chart.set_axis_range(Axis.LEFT, 0, 6)
+    chart.set_axis_labels(Axis.BOTTOM, ["Generation"])
+    chart.set_axis_labels(Axis.LEFT, ["Lift"])
+    chart.set_axis_positions(2, [50])
+    chart.set_axis_positions(3, [50])
+    chart.add_data(range(len(avgFitnessHistory)))
+    chart.add_data(minFitnessHistory)
+    chart.add_data(range(len(avgFitnessHistory)))
+    chart.add_data(avgFitnessHistory)
+    chart.add_data(range(len(avgFitnessHistory)))
+    chart.add_data(maxFitnessHistory)
+    chart.add_data(range(len(testFitnessHistory)))
+    chart.add_data(testFitnessHistory)
+    
+    open("chart.html", "w").write("<img src=\"%s\"/>" % chart.get_url())

### 2010-04-16 ann.py
Undo accidental commit of everything
git-svn-id: http://svn.scesoc.ca/evolution/trunk@67 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@ -12,58 +12,8 @@ import math
 
 log = logging.getLogger("ann")
 
-class Parameters(ctypes.Structure):
-  _fields_ = [("ih", 4 * (19 * ctypes.c_float)),
-              ("c", 4 * (19 * ctypes.c_float)),
-              ("w", 4 * ctypes.c_float),
-              ("ho", 4 * ctypes.c_float)]
-
-  def _float_list_str(self, l):
-    return ", ".join("%4.01g" % el for el in l)
-
-  def __str__(self):
-    s = []
-    s.append("Parameters(\n")
-    for i in range(4):
-      ihl = list(self.ih[i])
-      s.append("  ih[%d]=[0: %s,\n" % (i, self._float_list_str(ihl[0:10])))
-      s.append("        10: %s]\n" % (self._float_list_str(ihl[10:])))
-    for i in range(4):
-      cl = list(self.c[i])
-      s.append("  c[%d] =[0: %s,\n" % (i, self._float_list_str(cl[0:10])))
-      s.append("        10: %s]\n" % (self._float_list_str(cl[10:])))
-    s.append("  w =%s,\n" % self._float_list_str(list(self.w)))
-    s.append("  ho=%s,\n" % self._float_list_str(list(self.ho)))
-    s.append(")")
-    return "".join(s)
-
-  def _array_repr(self, dims, value):
-    if dims == []:
-      return "%e" % value
-
-    typestr = "ctypes.c_float"
-    for dim in reversed(dims):
-      typestr = "(%d*%s)" % (dim, typestr)
-    
-    valuestr = typestr + "(" + ", ".join(self._array_repr(dims[1:], subvalue) for subvalue in value) + ")"
-    return valuestr
-
-  def __repr__(self):
-    r = ["Parameters(\n"]
-    r.append("  ih=%s,\n" % self._array_repr([4, 19], self.ih))
-    r.append("  c=%s,\n" % self._array_repr([4, 19], self.c))
-    r.append("  w=%s,\n" % self._array_repr([4], self.w))
-    r.append("  ho=%s\n" % self._array_repr([4], self.ho))
-    r.append(")")
-    return "".join(r)
-
-  @staticmethod
-  def from_file(annfile):
-    return eval(compile(open(annfile).read(), annfile, "eval"),
-                {"Parameters": Parameters,
-                 "ctypes": ctypes})
-
[…]

### 2010-04-16 ann_kernels.cu

@@ -2,12 +2,13 @@
 #include <math_constants.h>
 
 #define ARRAY_SIZE(array) (sizeof(array) / sizeof((array)[0]))
+#define NODES_PER_LAYER 6
 
 struct Parameters {
-  float ih[4][19];  // input->hidden edge weight
-  float c[4][19];   // RBF center
-  float w[4];       // RBF width
-  float ho[4];      // hidden->output edge weight
+  float ih[NODES_PER_LAYER][19];  // input->hidden edge weight
+  float c[NODES_PER_LAYER][19];   // RBF center
+  float w[NODES_PER_LAYER];       // RBF width
+  float ho[NODES_PER_LAYER];      // hidden->output edge weight
 };
 
 __global__ void evaluate(
@@ -35,7 +36,7 @@ __global__ void evaluate(
     float output = 0.f;
     
     // TODO: Make sure this gets unrolled
-    for (int j = 0; j < 4; j++) {
+    for (int j = 0; j < NODES_PER_LAYER; j++) {
       float d2 = 0.f; // input to hidden node j
       for (int i = 0; i < 19; i++) {
         const float d = inputs[i] * p->ih[j][i] - p->c[j][i];

### 2010-04-16 blind.py

@@ -21,10 +21,9 @@ def main(annfile):
   for rank0, (index, output) in enumerate(sortedOutputs):
     indexToRank[index] = rank0 + 1
 
-  originalBlindSet = open("test.tsv")
-  for index, (row, output) in enumerate(zip(originalBlindSet, outputs)):
+  for index, output in enumerate(outputs):
     rank = indexToRank[index]
-    print "%s\t%d" % (row.strip(), rank)
+    print index, output, rank
 
 if __name__ == "__main__":
   if len(sys.argv) != 2:

### 2010-04-16 genops.py

@@ -21,7 +21,7 @@ Population generation / initializer.  Uses method from Montana and Davis.
 def generatePop(generation):
     for i in range(POPSIZE):
         nextMember = Parameters()
-        for j in range(4):
+        for j in range(ANN.NODES_PER_LAYER):
             for k in range(19):
                 nextMember.ih[j][k] = getInitialFloat()
                 nextMember.c[j][k] = getInitialFloat()
@@ -39,14 +39,14 @@ Create a new generation, based on the current generation
 @type fitList: A list of float values
 @return: List of new parameters (IE, a new generation)
 """
-def generateGeneration(oldGen):
+def generateGeneration(oldGen, mutateValue):
     newGen = []
     newGen.append(oldGen[0])
     for i in range(POPSIZE - 1):
-        if random.choice([0,1]) == 1:
+        if 10*random.random() > mutateValue + 2.5:
             index = getIndex()
             newGen.append(
-                mutate(oldGen[index])
+                mutate(oldGen[index], mutateValue)
             )
         else:
             index = getIndex()
@@ -66,22 +66,22 @@ Mutation operator.  Uses MUTATE NODES operator from Montana and Davis.
 @type xman: Parameters (see class Parameters, above)
 @return Mutated Parameters object
 """
-def mutate(xman):  
+def mutate(xman, mutateValue):  
     xmanJr = Parameters()
 
-    for i in range(4):
+    for i in range(ANN.NODES_PER_LAYER):
         for j in range(19):
             xmanJr.ih[i][j] = xman.ih[i][j]
             xmanJr.c[i][j] = xman.c[i][j]
         xmanJr.w[i] = xman.w[i]
         xmanJr.ho[i] = xman.ho[i]
 
-    node = random.randint(0,3)
+    node = random.randint(0,ANN.NODES_PER_LAYER-1)
     for i in range(19):
-        xmanJr.ih[node][i] += getMutationValue()
-        xmanJr.c[node][i] += getMutationValue()
-    xmanJr.w[node] += getMutationValue()
-    xmanJr.ho[node] += getMutationValue()
+        xmanJr.ih[node][i] += getMutationValue(mutateValue)
+        xmanJr.c[node][i] += getMutationValue(mutateValue)
+    xmanJr.w[node] += getMutationValue(mutateValue)
+    xmanJr.ho[node] += getMutationValue(mutateValue)
     
     return xmanJr
 
[…]

### 2010-04-16 genplot.py

@@ -1,51 +1,10 @@
-from pygooglechart import XYLineChart, Axis
-import cgi
+import matplotlib.pyplot as plt
 import numpy
 
-minFitnessHistory = []
-avgFitnessHistory = []
-maxFitnessHistory = []
-testFitnessHistory = []
-
 def plot():
-    pass
+    plt.show()
 
-def addGeneration(fitnessList, testFitness, genNumber):
-    fmin = min(fitnessList)
-    avg = sum(fitnessList) / len(fitnessList)
-    fmax = max(fitnessList)
-    print "G%d: avg %.02f min %.02f max %.02f test %.02f" % (
-        genNumber,
-        avg,
-        fmin,
-        fmax,
-        testFitness
-    )
-
-    minFitnessHistory.append(fmin)
-    avgFitnessHistory.append(avg)
-    maxFitnessHistory.append(fmax)
-    testFitnessHistory.append(testFitness)
-    
-    #xMax = len(avgFitnessHistory) * 120/100
-    xMax = 120
-    
-    chart = XYLineChart(640, 400, x_range=(0, xMax), y_range=(0, 6),
-      legend=["min", "avg", "max", "test"],
-      colours=["66CC00", "FF9933", "FF0000", "0000FF"])
-    chart.set_axis_range(Axis.BOTTOM, 0, xMax)
-    chart.set_axis_range(Axis.LEFT, 0, 6)
-    chart.set_axis_labels(Axis.BOTTOM, ["Generation"])
-    chart.set_axis_labels(Axis.LEFT, ["Lift"])
-    chart.set_axis_positions(2, [50])
-    chart.set_axis_positions(3, [50])
-    chart.add_data(range(len(avgFitnessHistory)))
-    chart.add_data(minFitnessHistory)
-    chart.add_data(range(len(avgFitnessHistory)))
-    chart.add_data(avgFitnessHistory)
-    chart.add_data(range(len(avgFitnessHistory)))
-    chart.add_data(maxFitnessHistory)
-    chart.add_data(range(len(testFitnessHistory)))
-    chart.add_data(testFitnessHistory)
-    
-    open("chart.html", "w").write("<img src=\"%s\"/>" % chart.get_url())
+def addGeneration(fitnessList, testFit, genNumber):
+    plt.scatter(genNumber, max(fitnessList))
+    plt.scatter(genNumber, testFit, marker="x")
+    plt.scatter(genNumber, min(fitnessList))

### 2010-04-16 input.py

@@ -111,11 +111,7 @@ class BlindSet(object):
 class Input(object):
   def __init__(self, infile, rand, blindfile=None):
     reader = csv.reader(map(string.strip, open(infile)), delimiter="\t")
-    self.data = list(reader)
-    #self._bayes(self.data, 2)
-    #self._bayes(self.data, 5)
-    self.data = [map(float, row) for row in self.data]
-
+    self.data = [map(float, row) for row in reader]
     featureCount = len(self.data[0]) - 1
 
     self._initIndexes()
@@ -140,8 +136,8 @@ class Input(object):
     padding = [0.0] * (19 - len(self.data[0]))
     self.data = [row + padding for row in self.data]
     
-    # Numerically, ANNs behave much more nicely with normalized data
-    quantiles = self._estimateQuantiles(self.data)
+    dataSet = DataSet(self.data[:numPositives], self.data[numPositives:])
+    self.trainSet, self.testSet = dataSet.split(70, rand)
     
     if blindfile is not None:
       blindreader = csv.reader(map(string.strip, open(blindfile)), delimiter="\t")
@@ -154,119 +150,7 @@ class Input(object):
       blinddata = [map(row.__getitem__, self._indexes) for row in blinddata]
       blinddata = [row + padding for row in blinddata]
 
-      self._normalize(blinddata, quantiles)
-
       self.blindSet = BlindSet(blinddata)
-
-    self._normalize(self.data, quantiles)
-
-    dataSet = DataSet(self.data[:numPositives], self.data[numPositives:])
-    self.trainSet, self.testSet = dataSet.split(70, rand)
-
-  def _bayes(self, data, featureIndex):
-    pos = neg = 0
-    p = {}
-    n = {}
-    for row in data:
-      v, Class = row[featureIndex], row[-1]
-
-      if Class == "1":
-        pos += 1
-        if v not in p:
-          p[v] = 0
-        p[v] += 1
-      else:
-        neg += 1
-        if v not in n:
-          n[v] = 0
-        n[v] += 1
-    
-    llrs = {} # "log likelihood ratios"
-    import sys, math
-    sys.stderr.write("pos: %d neg: %d ratio: %.03f\n" %
-      (pos, neg, float(pos)/float(neg))
-    )
-    allvalues = set(p.keys() + n.keys())
-    priorp = 1.0 / len(allvalues)
-    smoothing = 1.0
[…]

### 2010-04-16 input.py
Add normalization to Input class
git-svn-id: http://svn.scesoc.ca/evolution/trunk@68 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@ -111,7 +111,11 @@ class BlindSet(object):
 class Input(object):
   def __init__(self, infile, rand, blindfile=None):
     reader = csv.reader(map(string.strip, open(infile)), delimiter="\t")
-    self.data = [map(float, row) for row in reader]
+    self.data = list(reader)
+    #self._bayes(self.data, 2)
+    #self._bayes(self.data, 5)
+    self.data = [map(float, row) for row in self.data]
+
     featureCount = len(self.data[0]) - 1
 
     self._initIndexes()
@@ -136,8 +140,8 @@ class Input(object):
     padding = [0.0] * (19 - len(self.data[0]))
     self.data = [row + padding for row in self.data]
     
-    dataSet = DataSet(self.data[:numPositives], self.data[numPositives:])
-    self.trainSet, self.testSet = dataSet.split(70, rand)
+    # Numerically, ANNs behave much more nicely with normalized data
+    quantiles = self._estimateQuantiles(self.data)
     
     if blindfile is not None:
       blindreader = csv.reader(map(string.strip, open(blindfile)), delimiter="\t")
@@ -150,7 +154,119 @@ class Input(object):
       blinddata = [map(row.__getitem__, self._indexes) for row in blinddata]
       blinddata = [row + padding for row in blinddata]
 
+      self._normalize(blinddata, quantiles)
+
       self.blindSet = BlindSet(blinddata)
+
+    self._normalize(self.data, quantiles)
+
+    dataSet = DataSet(self.data[:numPositives], self.data[numPositives:])
+    self.trainSet, self.testSet = dataSet.split(70, rand)
+
+  def _bayes(self, data, featureIndex):
+    pos = neg = 0
+    p = {}
+    n = {}
+    for row in data:
+      v, Class = row[featureIndex], row[-1]
+
+      if Class == "1":
+        pos += 1
+        if v not in p:
+          p[v] = 0
+        p[v] += 1
+      else:
+        neg += 1
+        if v not in n:
+          n[v] = 0
+        n[v] += 1
+    
+    llrs = {} # "log likelihood ratios"
+    import sys, math
+    sys.stderr.write("pos: %d neg: %d ratio: %.03f\n" %
+      (pos, neg, float(pos)/float(neg))
+    )
+    allvalues = set(p.keys() + n.keys())
+    priorp = 1.0 / len(allvalues)
+    smoothing = 1.0
[…]

### 2010-04-16 ann.py
Fix random seed naming
git-svn-id: http://svn.scesoc.ca/evolution/trunk@69 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@ -400,7 +400,7 @@ def main():
   logging.basicConfig(level=logging.DEBUG)
   np.set_printoptions(precision=3, edgeitems=3, threshold=20)
 
-  randSample = random.Random(input.RAND_SAMPLE)
+  randSample = random.Random(input.SAMPLE_SEED)
   
   a = ANN()
   inp = input.Input("train3.tsv", randSample)

### 2010-04-16 ann_kernels.cu
Fix number of nodes per layer
git-svn-id: http://svn.scesoc.ca/evolution/trunk@70 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@ -2,7 +2,7 @@
 #include <math_constants.h>
 
 #define ARRAY_SIZE(array) (sizeof(array) / sizeof((array)[0]))
-#define NODES_PER_LAYER 6
+#define NODES_PER_LAYER 4
 
 struct Parameters {
   float ih[NODES_PER_LAYER][19];  // input->hidden edge weight

### 2010-04-16 ann_cpu.py
Use population size that gives reasonable execution time for CPU implementation; use test network parameters that give interesting outputs for the normalized input data
git-svn-id: http://svn.scesoc.ca/evolution/trunk@71 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@ -1,8 +1,4 @@
 #!/usr/bin/python
-import pycuda.autoinit
-from pycuda import driver
-from pycuda import compiler
-from pycuda import tools
 from itertools import islice
 import numpy as np
 import ctypes
@@ -14,19 +10,10 @@ log = logging.getLogger("ann")
 
 class ANN(object):
   NODES_PER_LAYER = 4
-  mod = compiler.SourceModule(open("ann_kernels.cu").read())
 
   def prepare(self, trainSet, popSize):
-    """Prepare for many parallel ANN fitness calculations.
+    """Prepare for many ANN fitness calculations.
     
-                         len(trainSet[0]) x len(trainSet)
-                       <--     training instances    -->
-                       +--------+--------+--------+----+
-       network       p0|blk(0,0)|blk(1,0)|blk(2,0)| .. |
-       params        p1|blk(0,1)|blk(1,1)|   ..   | .. |
-      (popSize x     p2|blk(0,2)|   ..   |   ..   | .. |
-    sizeof(Params))    +--------+--------+--------+----+
-     
     @param trainSet: training set
     @type trainSet: input.DataSet
     @param popSize: number of networks which will be evaluated in each run
@@ -37,89 +24,15 @@ class ANN(object):
     log.debug("training set size: %d", self.trainSet.size)
     log.debug("population size: %d", self.popSize)
 
-    # Calculate block/grid size and prepare evaluate() kernel.
-    # (avoids Function.__call__ overhead)
-    maxBlockDimX = driver.Context.get_device().get_attribute(
-      driver.device_attribute.MAX_BLOCK_DIM_X
-    )
-    self.evaluateBlockDim = (maxBlockDimX, 1, 1)
-    log.debug("evaluate kernel block dim: %r", self.evaluateBlockDim)
-    
-    blockDimX = self.evaluateBlockDim[0]
-    self.evaluateGridDim = ((self.trainSet.size + blockDimX - 1) / blockDimX,
-                            self.popSize)
-    log.debug("evaluate kernel grid dim: %r", self.evaluateGridDim)
-
-    self.evaluateKernel = self.mod.get_function("evaluate")
-    self.evaluateKernel.prepare(
-      (np.intp, np.uint32, np.intp, np.uint32, np.intp),
-      block=self.evaluateBlockDim
-    )
-
-    # Calculate block/grid size and prepare nlargest() kernel.
[…]

### 2010-04-16 bench.py
Add ANN benchmarking code
git-svn-id: http://svn.scesoc.ca/evolution/trunk@72 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@ -11,9 +11,9 @@ def timefun(fun, *args, **kwargs):
 
   deltas = []
   for _ in range(repeat):
-    start = time.clock()
+    start = time.time()
     ret = fun(*args)
-    end = time.clock()
+    end = time.time()
     deltas.append(end - start)
   
   avg = sum(deltas)/len(deltas)

### 2010-04-16 bench_ann.py

@@ -1,0 +1,87 @@
+#!/usr/bin/python -u
+"""Benchmark ANN evaluate, nlargest and lift calculation.
+
+This program takes command-line arguments to control the benchmarking.
+"""
+import optparse
+import ann_cpu
+import ann
+import bench
+import input
+import itertools
+import sys
+import random
+
+def initializeData():
+  global INPUT_SET, ANN_PARAM
+  rand = random.Random(input.SAMPLE_SEED)
+  inp = input.Input("train3.tsv", rand)
+  INPUT_SET = inp.trainSet
+  
+  # This set of parameters gives a good variation of network output values for
+  # normalized train3.tsv, so it exercises nlargest with representative data.
+  p = ann.Parameters()
+  for i in range(19):
+    p.ih[0][i] = p.ih[1][i] = p.ih[2][i] = p.ih[3][i] = 0.0
+    p.c[0][i] = p.c[1][i] = p.c[2][i] = p.c[3][i] = 0.0
+
+  p.ih[0][0] = 1.0
+
+  p.w[0] = -1.0
+  p.w[1] = p.w[2] = p.w[3] = 0.0
+
+  p.ho[0] = 1.0
+  p.ho[1] = p.ho[2] = p.ho[3] = 0.0
+  
+  ANN_PARAM = p
+
+def generatePopulation(size):
+  return [ANN_PARAM] * size
+
+def _repeatInstances(instances, size):
+  repeated = instances * ((size + len(instances) - 1) / len(instances))
+  return repeated[0:size]
+ 
+def generateTrainSet(size):
+  numPositives = len(INPUT_SET.positives) * size / INPUT_SET.size
+  return input.DataSet(
+    _repeatInstances(INPUT_SET.positives, numPositives),
+    _repeatInstances(INPUT_SET.negatives, size - numPositives)
+  )
+
+def main(popSizes, trainSizes):
+  initializeData()
+  
+  sys.stdout.write("%12s%12s%12s%12s\n" % (
+    "PopSize", "TrainSize", "CPU (ms)", "GPU (ms)"
+  ))
+  for trainSize in trainSizes:
+    # generateTrainSet is more expensive than generatePopulation
+    trainSet = generateTrainSet(trainSize)
+    for popSize in popSizes:
+      population = generatePopulation(popSize)
+      n = trainSet.size * 20 / 100
+      
+      sys.stdout.write("%12d%12d" % (popSize, trainSize))
+      
+      def benchImpl(impl):
[…]

### 2010-04-17 doc/paper.tex
Add experimental results.
git-svn-id: http://svn.scesoc.ca/evolution/trunk@73 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 % ----------------------------------------------------------------------------
-Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this
+To evaluate the performance of our system, we compared execution time of equivalent algorithms running on commodity desktop CPUs and GPUs.
+Each series represents one type of experimental hardware (Table~\ref{tab:experimental-hardware}).
+Each data point is an average over 10 runs of the algorithm with the same parameters.
+
+\begin{table}
+	\centering
+	\begin{tabular}{ll}
+	\textbf{Legend Label} & \textbf{Hardware} \\
+	\hline
+	Core2 & Intel Core 2 Q9450 CPU at 2.66 GHz \\
+	GTX275 & NVidia GTX275 GPU \\
+	\hline
+	\end{tabular}
+	\caption{Experimental hardware}
+	\label{tab:experimental-hardware}
+\end{table}
 
@@
 % ----------------------------------------------------------------------------
-Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this
+In most cases, the GPU implementation resulted in approximately an order of magnitude speedup over a CPU implementation (Figure~\ref{fig:training-performance}).
+For low population sizes, GPU initialization and data copying dominates.
+
+\begin{figure}
+	\includegraphics[width=\textwidth]{fig-performance}
+	\caption{Training performance as a function of population size (left) and training set size (right). Note logarithmic axes.}
+	\label{fig:training-performance}
+\end{figure}
 

### 2010-04-17 doc/paper.tex
Write abstract, cut out too verbose Introduction, move same content more concisely into Background
git-svn-id: http://svn.scesoc.ca/evolution/trunk@74 fee98b4c-d1c3-4d34-9525-a468159e4a48

--- 2010-04-15 doc/paper.tex
@@
 \begin{abstract}
-Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this
+We present an implementation of genetic algorithm (GA) training of artificial neural networks (ANNs) targeting commodity graphics cards (GPUs).
+By carefully mapping the problem onto the unique GPU architecture, we achieve order-of-magnitude speedup over a conventional CPU implementation.
+Furthermore, we show that the speedup is consistent across a wide range of population and training set sizes.
+This performance boost enables classifier design to span a wider range of population sizes and number of generations affording better classifiers and improved understanding of long-term trends in genetic algorithm convergence.
+Finally, we apply this method to the design of a classifier for the 2009 UC San Diego Data Mining Contest.
 \end{abstract}
@@
 % ############################################################################
-Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this
 
@@
 % ----------------------------------------------------------------------------
-Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this
+Artificial neural networks (ANNs) are very flexible classifiers.
+By varying the network structure and parameters thereof, they can represent many aspects of the function to be approximated, including nonlinearities and high-order terms.
+However, the complexity of these classifiers implies that finding optimal parameters coincides with a search in a very high-dimensional parametric space.
+This is known as an optimization problem, for which a wide variety of solutions are known.
+
+Greedy approaches to ANN training include gradient descent, Gauss-Newton and Levenberg–Marquardt, which are first order, second order and hybrid methods, respectively.
+These methods seek extrema using the local behaviour of the error function and are thus susceptible to getting stuck at local extrema rather than finding the global extremum.
+
+Stochastic approaches offer the promise of "eventually" reaching the global extremum by asymptotically exploring the entire search space.
+Simulated annealing, for example, iteratively generates candidate solutions by randomly perturbing the previous solution.
+Early in the process, we inject a high amount of randomness to find global trends in the error function, while later in the process we inject less randomness to find the apex of the (hopefully) global extremum.
 
@@
 % ----------------------------------------------------------------------------
-Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this
+Genetic algorithms (GAs) rely on multiple candidate solutions to simultaneously explore the search space.
+The "individuals" are randomly mutated, mated and selected during each of several generations.
+Mutation and mating is done using problem-specific representation of candidate solutions and "genetic operators", while selection is governed by a fitness function (classifier performance).
+The competitive bias imposed by selection attempts to mimic the "survival of the fittest" principle so often seen in Nature.
+
+GA training of ANNs is very computationally-intensive.
+For each generation ($10^1$--$10^2$) individuals), the fitness of each individual is calculated by applying each neural network to the entire training set ($10^2$--$10^6$ instances) and taking the classification accuracy for that network.
+Each ANN evaluation itself requires a large number of multiplications (which increases with the number of features and hidden nodes) in addition to evaluation of transcendentals for the most commonly-used node activation function.
+Thus, a CPU implementation of GA training of ANNs for any but the smallest networks and training sets is infeasible.
 
@@
 % ----------------------------------------------------------------------------
-Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this
+
+Compute Unified Device Architecture (CUDA) is a platform which enables the use of NVIDIA graphics cards (GPUs) for general purpose programming, including scientific computations.
+GPUs have recently become quite affordable to the point that some new notebook computers now include built-in GPUs.
+Because of the nature of graphics processing, these are massively parallel processors optimized for memory bandwidth and floating-point computation.
+This is a very good match for our GA-ANN training problem.
+
+The CUDA programming model is designed to allow algorithms to scale with the GPU's parallelism parameters.
+This allows the same algorithm to run on different present-day hardware as well as future hardware, with minimal changes.
+However, to accomplish this, the programming model is different in many ways from the conventional CPU programming.
+In particular, to best leverage GPU hardware, CUDA code must be carefully written to balance mathematical operations with memory accesses and to have certain specific optimal memory access patterns.
+These are both well-known problems in high performance computing, but they are further exacerbated by the design of the CUDA architecture.
 

### 2010-04-17 doc/paper.tex
Shorten some text and add info about data mining contest (in particular lift@20)
git-svn-id: http://svn.scesoc.ca/evolution/trunk@75 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 By varying the network structure and parameters thereof, they can represent many aspects of the function to be approximated, including nonlinearities and high-order terms.
-However, the complexity of these classifiers implies that finding optimal parameters coincides with a search in a very high-dimensional parametric space.
+However, the complexity of these classifiers implies that finding optimal parameters is a search in a very high-dimensional parameter space.
 This is known as an optimization problem, for which a wide variety of solutions are known.
 
-Greedy approaches to ANN training include gradient descent, Gauss-Newton and Levenberg–Marquardt, which are first order, second order and hybrid methods, respectively.
-These methods seek extrema using the local behaviour of the error function and are thus susceptible to getting stuck at local extrema rather than finding the global extremum.
-
-Stochastic approaches offer the promise of "eventually" reaching the global extremum by asymptotically exploring the entire search space.
-Simulated annealing, for example, iteratively generates candidate solutions by randomly perturbing the previous solution.
-Early in the process, we inject a high amount of randomness to find global trends in the error function, while later in the process we inject less randomness to find the apex of the (hopefully) global extremum.
+Some approaches to ANN training include gradient descent, Gauss-Newton and Levenberg–Marquardt.
+Greedy approaches such as these seek extrema using the local behaviour of the error function and are thus susceptible to getting stuck at local extrema rather than finding the global extremum.
+Stochastic approaches asymptotically explore the entire search space, offering the promise of "eventually" reaching the global extremum.
 
@@
 The "individuals" are randomly mutated, mated and selected during each of several generations.
-Mutation and mating is done using problem-specific representation of candidate solutions and "genetic operators", while selection is governed by a fitness function (classifier performance).
+Mutation and mating is done using a problem-specific representation of candidate solutions and "genetic operators." Selection is governed by a fitness function (classifier performance).
 The competitive bias imposed by selection attempts to mimic the "survival of the fittest" principle so often seen in Nature.
@@
 For each generation ($10^1$--$10^2$) individuals), the fitness of each individual is calculated by applying each neural network to the entire training set ($10^2$--$10^6$ instances) and taking the classification accuracy for that network.
-Each ANN evaluation itself requires a large number of multiplications (which increases with the number of features and hidden nodes) in addition to evaluation of transcendentals for the most commonly-used node activation function.
+Each ANN evaluation itself requires a large number of multiplications (which increases with the number of features and hidden nodes) in addition to evaluation of transcendentals for the more common node activation functions.
 Thus, a CPU implementation of GA training of ANNs for any but the smallest networks and training sets is infeasible.
@@
 % ----------------------------------------------------------------------------
+NVidia's Compute Unified Device Architecture (CUDA) is a programming paradigm for massively parallel processors present in off-the-shelf graphics cards (GPUs).
+GPUs consist of many independent floating-point computation units, providing significant speedup to data-parallel compute-intensive applications.
+Each unit is connected to on-board memory using a very wide bus, enabling very high memory bandwidth provided certain particular memory access patterns are respected.
+The combination of these features makes CUDA an ideal platform for GA-ANN training.
 
-Compute Unified Device Architecture (CUDA) is a platform which enables the use of NVIDIA graphics cards (GPUs) for general purpose programming, including scientific computations.
-GPUs have recently become quite affordable to the point that some new notebook computers now include built-in GPUs.
-Because of the nature of graphics processing, these are massively parallel processors optimized for memory bandwidth and floating-point computation.
-This is a very good match for our GA-ANN training problem.
-
-The CUDA programming model is designed to allow algorithms to scale with the GPU's parallelism parameters.
-This allows the same algorithm to run on different present-day hardware as well as future hardware, with minimal changes.
-However, to accomplish this, the programming model is different in many ways from the conventional CPU programming.
-In particular, to best leverage GPU hardware, CUDA code must be carefully written to balance mathematical operations with memory accesses and to have certain specific optimal memory access patterns.
-These are both well-known problems in high performance computing, but they are further exacerbated by the design of the CUDA architecture.
+% ----------------------------------------------------------------------------
+\subsection{Data Mining Contest} \label{contest}
+% ----------------------------------------------------------------------------
+The data for our classifier was provided by the 2009 UC San Diego Data Mining Contest "E-commerce Transaction Anomaly Classification" and consist of 94682 instances of 19 features with a 1:50 class imbalance.
+The evaluation metric was "lift at 20\%", which can be understood as the ratio of the true positive rate in the top 20\% of instances ranked by the classifier to the overall positive rate of the original data set.
+This metric was directly used as the fitness of each classifier being evaluated by the GA.
 

### 2010-04-17 doc/paper.tex
Add info about CUDA architecture and algorithm design (nlargest and count)
git-svn-id: http://svn.scesoc.ca/evolution/trunk@76 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 
+The computational unit in CUDA is the "kernel", which is a C function executed in parallel on the grid of available compute units in the CUDA device.
+The function can request its position in this grid such that it can apply the same algorithm to values at different addresses in memory.
+In addition, if groups of kernels access sequential addresses in memory, the memory fetches are done in parallel, maximizing bus bandwidth utilization.
+This is referred to as coalesing.
+
+The CUDA programming model defines global and shared memory.
+Global memory is on-board but off-chip, expensive for random access and plentiful (100s of MB).
+Shared memory is on-chip, very fast for random access but very limited (16 KB).
+
 % ----------------------------------------------------------------------------
@@
 This metric was directly used as the fitness of each classifier being evaluated by the GA.
+ANN outputs are treated as confidence values and used for ranking.
 
@@
 % ############################################################################
-Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this
+The training process can be divided in several parts which are repeated for each generation until sufficient classifier performance is obtained.
+\begin{enumerate}
+	\item Genetic algorithm processing -- initialization, mating, mutation, selection
+	\item Computation of ANN outputs
+	\item Calculation of top 20\% threshold value (each individual)
+	\item Counting of the positive instances with outputs above this threshold (each individual)
+\end{enumerate}
+
+Item 1 is not very computationally intensive and was not parallelized.
+Items 2, 3 and 4 were all implemented on the CUDA platform.
+Performance results are given for the combined execution time of these three items.
 
@@
 Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this
+
+% ----------------------------------------------------------------------------
+\subsection{Calculation of Top 20\% Threshold} \label{implementation}
+% ----------------------------------------------------------------------------
+To efficiently calculate the top 20\% threshold in a list of $n$ ANN output values, we can insert all values into a $k$-sized minheap where $k = 0.20 * n$, and take the root of the heap as the threshold.
+Each minheap insertion costs $O(\log k)$ time.
+Because device shared memory (16~KB) is not large enough for our value of $k$, we use $p$ passes of $\frac{k}{p}$-sized heaps with which we find the top 5\% threshold, then the top 10\%, etc.
+The overall complexity of our algorithm is $O(np \log \frac{k}{p})$.
+Parallelization is across individuals.
+
+% ----------------------------------------------------------------------------
+\subsection{Counting of Positives Above Threshold} \label{implementation}
+% ----------------------------------------------------------------------------
+By keeping positive instances first in our training set, we avoid the need to consult the class of each instance (saving a costly global memory read).
+We simply iterate over the first $numPositives$ ANN outputs and count the outputs over the individual's 20\% threshold.
+Parallelization is across both individuals and groups of outputs.
+For each individual, we keep m counters which accumulate counts for indexes $i$ where $i \equiv 0 \bmod{m}$, $i \equiv 1 \bmod{m}$, ..., $i \equiv m - 1 \bmod{m}$.
+This enables parallelization even when there are not enough individuals in the population to fully occupy the device.
+Each batch of $m$ kernels accesses $m$ sequential locations in memory, which results in coalesced reads.
 

### 2010-04-18 doc/paper.tex
Slightly change wording in abstract
git-svn-id: http://svn.scesoc.ca/evolution/trunk@77 fee98b4c-d1c3-4d34-9525-a468159e4a48
Add Computation of ANN Output Values and Conclusion
git-svn-id: http://svn.scesoc.ca/evolution/trunk@78 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 This performance boost enables classifier design to span a wider range of population sizes and number of generations affording better classifiers and improved understanding of long-term trends in genetic algorithm convergence.
-Finally, we apply this method to the design of a classifier for the 2009 UC San Diego Data Mining Contest.
+Finally, we demonstrate this method in the context of the 2009 UC San Diego Data Mining Contest.
 \end{abstract}
@@
 % ----------------------------------------------------------------------------
+\subsection{Computation of ANN Output Values} \label{implementation}
+% ----------------------------------------------------------------------------
+ANN output values are computed by reading feature values and applying the appropriate operations represented by the network topology and parameters.
+Feature data are organized to enable memory read coalescing.
+This part of the process is particularly well-suited to the CUDA architecture because it has a very high ratio of mathematical operations to memory accesses (160 multiplications, 4 exponentiations and 80 bytes of memory accesses for our ANN topology).
+
+% ----------------------------------------------------------------------------
 \subsection{Calculation of Top 20\% Threshold} \label{implementation}
@@
 % ############################################################################
-Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this
+We have presented an implementation of GA training of ANN classifiers for the CUDA platform for GPU programming.
+By carefully designing memory organization, algorithm computational load and memory access patterns, we have obtained a 10-fold speedup compared to a conventional sequential CPU implementation.
+Additional workaround were necessary for the low amount of fast on-chip GPU memory.
+Our method scales across population and training set sizes.
 

### 2010-04-18 doc/paper.tex
Tweak wording on how we gave perf numbers
git-svn-id: http://svn.scesoc.ca/evolution/trunk@79 fee98b4c-d1c3-4d34-9525-a468159e4a48
use en dash for two-author algorithm names
git-svn-id: http://svn.scesoc.ca/evolution/trunk@80 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 
-Some approaches to ANN training include gradient descent, Gauss-Newton and Levenberg–Marquardt.
+Some approaches to ANN training include gradient descent, Gauss--Newton and Levenberg-–Marquardt.
 Greedy approaches such as these seek extrema using the local behaviour of the error function and are thus susceptible to getting stuck at local extrema rather than finding the global extremum.
@@
 Items 2, 3 and 4 were all implemented on the CUDA platform.
-Performance results are given for the combined execution time of these three items.
+Performance results are given as the combined time for one generation of these last three items but exclude one-time initialization.
 

### 2010-04-18 doc/paper.tex
memory access patterns -> rules
git-svn-id: http://svn.scesoc.ca/evolution/trunk@81 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 GPUs consist of many independent floating-point computation units, providing significant speedup to data-parallel compute-intensive applications.
-Each unit is connected to on-board memory using a very wide bus, enabling very high memory bandwidth provided certain particular memory access patterns are respected.
+Each unit is connected to on-board memory using a very wide bus, enabling very high memory bandwidth provided certain particular memory access rules are respected.
 The combination of these features makes CUDA an ideal platform for GA-ANN training.

### 2010-04-18 doc/paper.tex
Remove redundant word
git-svn-id: http://svn.scesoc.ca/evolution/trunk@82 fee98b4c-d1c3-4d34-9525-a468159e4a48
Swap two paragraphs so we can use the "global memory" term later
git-svn-id: http://svn.scesoc.ca/evolution/trunk@83 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 NVidia's Compute Unified Device Architecture (CUDA) is a programming paradigm for massively parallel processors present in off-the-shelf graphics cards (GPUs).
-GPUs consist of many independent floating-point computation units, providing significant speedup to data-parallel compute-intensive applications.
+GPUs consist of many independent floating-point units, providing significant speedup to data-parallel compute-intensive applications.
 Each unit is connected to on-board memory using a very wide bus, enabling very high memory bandwidth provided certain particular memory access rules are respected.
 The combination of these features makes CUDA an ideal platform for GA-ANN training.
-
-The computational unit in CUDA is the "kernel", which is a C function executed in parallel on the grid of available compute units in the CUDA device.
-The function can request its position in this grid such that it can apply the same algorithm to values at different addresses in memory.
-In addition, if groups of kernels access sequential addresses in memory, the memory fetches are done in parallel, maximizing bus bandwidth utilization.
-This is referred to as coalesing.
 
@@
 Shared memory is on-chip, very fast for random access but very limited (16 KB).
+
+The computational unit in CUDA is the "kernel", which is a C function executed in parallel on the grid of available compute units in the CUDA device.
+The function can request its position in this grid such that it can apply the same algorithm to values at different addresses in memory.
+In addition, if groups of kernels access sequential addresses in global memory, the memory fetches are done in parallel, maximizing bus bandwidth utilization.
+This is referred to as memory access coalesing.
 

### 2010-04-18 doc/paper.tex
This abstract is too big!
git-svn-id: http://svn.scesoc.ca/evolution/trunk@84 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 % ############################################################################
-Catalin Patulea and Robert Peace\\
+Catalin Patulea, Robert Peace, and James Green\\
 School of Systems and Computer Engineering\\
@@
 Ottawa, Canada K1S 5B6\\
-{\em rpeace@sce.carleton.ca}
+{\em cat@vv.carleton.ca}, {\em rpeace@sce.carleton.ca}, {\em jrgreen@sce.carleton.ca}
 % ############################################################################
@@
 
-Some approaches to ANN training include gradient descent, Gauss--Newton and Levenberg-–Marquardt.
+Some approaches to ANN training include gradient descent, Gauss--Newton and Levenberg-Marquardt.
 Greedy approaches such as these seek extrema using the local behaviour of the error function and are thus susceptible to getting stuck at local extrema rather than finding the global extremum.
@@
 % ----------------------------------------------------------------------------
-Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this
 
@@
 % ----------------------------------------------------------------------------
-Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this
+We have implemented an algorithm which, given a tab seperated values file, converts nominal feature values in the file into numeric values.
+The algorithm replaces each nominal value in the file with the logarithm of the probability ratio between the positive and negative values observed given the nominal value.
+As a result, meaningful numeric values can be extracted from arbitrary nominal values without domain knowledge relating to the values.
 
@@
 % ----------------------------------------------------------------------------
-Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this
+In order to adapt to varying data sets, we have developped a flexible ANN model, allowing for an arbitrary number of hidden layers and nodes per layer (within the limits of GPU memory).
+All hidden layers must contain the same number of nodes, however input and output weights may be adjusted through the genetic algorithm so as to effectively reduce the number of nodes on a specific layer.
+
+The genetic mutation and crossover operations which we have implemented are based on the methods defined by Montana and Lawrence \cite{GA-ANN}.
+Of the operations presented in this paper, the MUTATE NODES and CROSSOVER WEIGHTS operations were determined to provide an optimal balance between performance and classifier accuracy, and have been implemented in our experiments.
+
+Preliminary analysis of test data has shown that a projection of the data onto a 2D space which maximize class separation produces four distinct eliptical regions.
+Therefore, test results have been obtained using an ANN which contains one layer of four radial basis function nodes.
 
@@
 % ----------------------------------------------------------------------------
-Filling this with text so that I can see what the document looks like when it is filled with text and to make sure that everything lines up nicely when it is full of text like this
+Montana and Lawrence \cite{GA-ANN} have demonstrated that a brief period of hill climbing after training with a genetic algorithm may increase the accuracy of classifier results.
+Thus, we could improve the accuracy of our methods by incorporating hill-climbing methods, such as the parallel backpropagation method described by Oei, Friedland and Janin \cite{backprop}.
 

### 2010-04-18 doc/paper.tex
Cut down on background, write introduction
git-svn-id: http://svn.scesoc.ca/evolution/trunk@85 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 % ############################################################################
+Genetic algorithms (GAs) are a stochastic, evolutionary approach to machine learning classifier training.
+While greedy methods can get stuck at local extrema, GAs  are theoretically capable of asymptotically reaching the global optimum.
+However, because they require the testing of a population of classifiers at each iteration, and the algorithm is repeated for many generations, they are extremely compute-intensive.
+This is exacerbated by the use of artificial neural networks (ANNs) as the classifier because ANNs themselves are very compute-intensive.
+Compared to a traditional sequential implementation of GA training of ANNs, the use of graphical processing units (GPUs) as parallel processors provides significant performance improvements, at a fraction of the price of alternatives such as cluster computing.
+GPUs, however, use a specialized programming paradigm which must be taken into account to leverage their full processing power.
+In this paper, we show an implementation of GA training of ANNs which achieves an order-of-magnitude speedup over a sequential algorithm.
 
@@
 % ----------------------------------------------------------------------------
-Artificial neural networks (ANNs) are very flexible classifiers.
-By varying the network structure and parameters thereof, they can represent many aspects of the function to be approximated, including nonlinearities and high-order terms.
-However, the complexity of these classifiers implies that finding optimal parameters is a search in a very high-dimensional parameter space.
-This is known as an optimization problem, for which a wide variety of solutions are known.
-
-Some approaches to ANN training include gradient descent, Gauss--Newton and Levenberg-Marquardt.
-Greedy approaches such as these seek extrema using the local behaviour of the error function and are thus susceptible to getting stuck at local extrema rather than finding the global extremum.
-Stochastic approaches asymptotically explore the entire search space, offering the promise of "eventually" reaching the global extremum.
+ANNs are composed of a network of nodes connected by weighted directed edges.
+In a radial basis function (RBF) network architecture, nodes are organized in an input layer, a hidden layers, and an output layer.
+Each layer is fully connected to the next.
+The nodes in the hidden layer calculate the distance between their inputs and a centre parameter, and pass a weighted distance to an RBF.
+The output node is a summer weighted by the incoming edges.
+The output is a single real value which can be interpreted as a classification confidence.
+ANN computation is dominated by the multiplications at the edges, proportional in number to the number of features (19) and the number of hidden nodes (4) and transcendental functions (exponentiation for Gaussian RBF) at the hidden nodes.
 
@@
 % ----------------------------------------------------------------------------
-Genetic algorithms (GAs) rely on multiple candidate solutions to simultaneously explore the search space.
+Genetic algorithms (GAs) use multiple candidate solutions to simultaneously explore the search space of ANN parameters.
 The "individuals" are randomly mutated, mated and selected during each of several generations.
-Mutation and mating is done using a problem-specific representation of candidate solutions and "genetic operators." Selection is governed by a fitness function (classifier performance).
+Mutation and mating is done using a problem-specific representation of candidate solutions and "genetic operators." Selection is performed by applying the classifier to the training set and calculating a fitness value (classifier performance).
 The competitive bias imposed by selection attempts to mimic the "survival of the fittest" principle so often seen in Nature.
 
-GA training of ANNs is very computationally-intensive.
-For each generation ($10^1$--$10^2$) individuals), the fitness of each individual is calculated by applying each neural network to the entire training set ($10^2$--$10^6$ instances) and taking the classification accuracy for that network.
-Each ANN evaluation itself requires a large number of multiplications (which increases with the number of features and hidden nodes) in addition to evaluation of transcendentals for the more common node activation functions.
-Thus, a CPU implementation of GA training of ANNs for any but the smallest networks and training sets is infeasible.
+The computation required for GA training of ANNs is proportional to the number of generations ($10^1$-$10^4$), to the size of each generation ($10^1$-$10^2$ individuals) and to the training set size ($10^2$-$10^6$ instances).
 
@@
 % ----------------------------------------------------------------------------
-NVidia's Compute Unified Device Architecture (CUDA) is a programming paradigm for massively parallel processors present in off-the-shelf graphics cards (GPUs).
+NVidia's Compute Unified Device Architecture (CUDA) is a programming paradigm for massively parallel GPUs found in off-the-shelf graphics cards.
 GPUs consist of many independent floating-point units, providing significant speedup to data-parallel compute-intensive applications.
-Each unit is connected to on-board memory using a very wide bus, enabling very high memory bandwidth provided certain particular memory access rules are respected.
+Each unit is connected to on-board memory using a very wide bus, enabling high memory bandwidth provided certain particular memory access rules are respected.
 The combination of these features makes CUDA an ideal platform for GA-ANN training.
@@
 The computational unit in CUDA is the "kernel", which is a C function executed in parallel on the grid of available compute units in the CUDA device.
-The function can request its position in this grid such that it can apply the same algorithm to values at different addresses in memory.
-In addition, if groups of kernels access sequential addresses in global memory, the memory fetches are done in parallel, maximizing bus bandwidth utilization.
-This is referred to as memory access coalesing.
+The function is aware of its position in this grid such that it can apply the same algorithm to data at different addresses in memory.
+In addition, if groups of kernels access sequential addresses in global memory, the memory fetches are combined into a single transaction, maximizing bus bandwidth utilization.
+This is referred to as memory access coalescing.
 
@@
 % ----------------------------------------------------------------------------
-The data for our classifier was provided by the 2009 UC San Diego Data Mining Contest "E-commerce Transaction Anomaly Classification" and consist of 94682 instances of 19 features with a 1:50 class imbalance.
-The evaluation metric was "lift at 20\%", which can be understood as the ratio of the true positive rate in the top 20\% of instances ranked by the classifier to the overall positive rate of the original data set.
-This metric was directly used as the fitness of each classifier being evaluated by the GA.
-ANN outputs are treated as confidence values and used for ranking.
+Our demonstration classifier is designed for the 2009 UC San Diego Data Mining Contest "E-commerce Transaction Anomaly Classification" and with training data thereof.
+The training data consist of 94682 instances of 19 mixed features with a 1:50 binary class imbalance.
+The evaluation metric is "lift at 20\%", which can be understood as the ratio of the true positive rate in the top 20\% ranked instances to the overall positive rate of the data set.
 
@@
 % ############################################################################
-The training process can be divided in several parts which are repeated for each generation until sufficient classifier performance is obtained.
+Each GA iteration consists of the following steps:
 \begin{enumerate}
-	\item Genetic algorithm processing -- initialization, mating, mutation, selection
 	\item Computation of ANN outputs
@@
 	\item Counting of the positive instances with outputs above this threshold (each individual)
+	\item Genetic algorithm processing -- mating, mutation, selection
 \end{enumerate}
 
-Item 1 is not very computationally intensive and was not parallelized.
-Items 2, 3 and 4 were all implemented on the CUDA platform.
+Items 1, 2 and 3 were all implemented on the CUDA platform.
 Performance results are given as the combined time for one generation of these last three items but exclude one-time initialization.
+Item 4 is not very computationally intensive and was not parallelized.
 
@@
 
+% Where does this fit?
+%  ANN outputs are treated as confidence values and used for ranking, and the lift of a classifier was directly used as the its fitness value for the GA.
+
 % ----------------------------------------------------------------------------
@@
 Each minheap insertion costs $O(\log k)$ time.
-Because device shared memory (16~KB) is not large enough for our value of $k$, we use $p$ passes of $\frac{k}{p}$-sized heaps with which we find the top 5\% threshold, then the top 10\%, etc.
+Because device shared memory (16~KB) is not large enough for our value of $k$, we instead use $p$ passes of $\frac{k}{p}$-sized heaps with which we find the top 5\% threshold, then the top 10\%, etc.
 The overall complexity of our algorithm is $O(np \log \frac{k}{p})$.

### 2010-04-18 doc/paper.tex
Better separate sentence
git-svn-id: http://svn.scesoc.ca/evolution/trunk@86 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 The output is a single real value which can be interpreted as a classification confidence.
-ANN computation is dominated by the multiplications at the edges, proportional in number to the number of features (19) and the number of hidden nodes (4) and transcendental functions (exponentiation for Gaussian RBF) at the hidden nodes.
+ANN computation is dominated by the multiplications at the edges, proportional in number to the number of features (19) and the number of hidden nodes (4), and transcendental functions (exponentiation for Gaussian RBF) at the hidden nodes.
 

### 2010-04-18 ann_kernels_twolayer.cu
!
git-svn-id: http://svn.scesoc.ca/evolution/trunk@88 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@ -6,12 +6,12 @@
 
 struct Parameters {
   float ih[NODES_PER_LAYER][19];			// input->hidden edge weight
-  float hh[NODES_PER_LAYER][NODES_PER_LAYER];// hidden->hidden edge weight
   float c[NODES_PER_LAYER][19];				// RBF center
-  float c2[NODES_PER_LAYER][NODES_PER_LAYER];// RBF center second layer
   float w[NODES_PER_LAYER];					// RBF width
   float w2[NODES_PER_LAYER];				// RBF width second layer
   float ho[NODES_PER_LAYER];				// hidden->output edge weight
+  float hh[NODES_PER_LAYER][NODES_PER_LAYER];// hidden->hidden edge weight
+  float c2[NODES_PER_LAYER][NODES_PER_LAYER];// RBF center second layer
 };
 
 __global__ void evaluate(

### 2010-04-18 bench_ann.py

@@ -82,6 +82,6 @@ def main(popSizes, trainSizes):
       sys.stdout.write("\n")
 
 if __name__ == "__main__":
-  #main([1, 10, 100, 1000, 10000], [1000])
-  main([100], [100, 1000, 10000, 100000])
+  main([1, 10, 100, 1000, 10000], [1000])
+  #main([100], [100, 1000, 10000, 100000])
 

### 2010-04-18 genops2layer.py

@@ -11,7 +11,7 @@ from classify import SampleTester
 log = logging.getLogger("genops")
 
 POPSIZE = 50
-TOTAL_GENS = 5000
+TOTAL_GENS = 10000
 
 """
 Population generation / initializer.  Uses method from Montana and Davis.
@@ -167,7 +167,7 @@ def main():
   logging.basicConfig(level=logging.INFO, stream=sys.stdout)
   np.set_printoptions(precision=3, edgeitems=3, threshold=20)
 
-  random.seed(5108) # used by the GA
+  random.seed(80085) # used by the GA
   randSample = random.Random(input.SAMPLE_SEED) # used for data set sampling
 
   inp = input.Input("train3-std.tsv", randSample)

### 2010-04-18 bench_ann.py
Use 10000 trainsize for popsize benchmark
git-svn-id: http://svn.scesoc.ca/evolution/trunk@89 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@ -82,6 +82,6 @@ def main(popSizes, trainSizes):
       sys.stdout.write("\n")
 
 if __name__ == "__main__":
-  main([1, 10, 100, 1000, 10000], [1000])
+  main([1, 10, 100, 1000, 10000], [10000])
   #main([100], [100, 1000, 10000, 100000])
 

### 2010-04-18 doc/paper.tex
CUDA is a platform, not a paradigm
git-svn-id: http://svn.scesoc.ca/evolution/trunk@87 fee98b4c-d1c3-4d34-9525-a468159e4a48
Add new perf data series and move GA/ANN process description to after data analysis
git-svn-id: http://svn.scesoc.ca/evolution/trunk@90 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 % ----------------------------------------------------------------------------
-NVidia's Compute Unified Device Architecture (CUDA) is a programming paradigm for massively parallel GPUs found in off-the-shelf graphics cards.
+NVidia's Compute Unified Device Architecture (CUDA) is a programming platform for massively parallel GPUs found in off-the-shelf graphics cards.
 GPUs consist of many independent floating-point units, providing significant speedup to data-parallel compute-intensive applications.
@@
 % ############################################################################
-Each GA iteration consists of the following steps:
-\begin{enumerate}
-	\item Computation of ANN outputs
-	\item Calculation of top 20\% threshold value (each individual)
-	\item Counting of the positive instances with outputs above this threshold (each individual)
-	\item Genetic algorithm processing -- mating, mutation, selection
-\end{enumerate}
-
-Items 1, 2 and 3 were all implemented on the CUDA platform.
-Performance results are given as the combined time for one generation of these last three items but exclude one-time initialization.
-Item 4 is not very computationally intensive and was not parallelized.
 
@@
 % ----------------------------------------------------------------------------
+Each GA iteration consists of the following steps:
+\begin{enumerate}
+	\item Computation of ANN outputs
+	\item Calculation of top 20\% threshold value (each individual)
+	\item Counting of the positive instances with outputs above this threshold (each individual)
+	\item Genetic algorithm processing -- mating, mutation, selection
+\end{enumerate}
+
+Items 1, 2 and 3 were all implemented on the CUDA platform.
+Performance results are given as the combined time for one generation of these last three items but exclude one-time initialization.
+Item 4 is not very computationally intensive and was not parallelized.
 
@@
 	\hline
-	Core2 & Intel Core 2 Q9450 CPU at 2.66 GHz \\
+	Core2 & Intel Core 2 Q9450 CPU \\
+	i7 & Intel i7 920 CPU \\
+	GTX260 & NVidia GTX260 GPU \\
 	GTX275 & NVidia GTX275 GPU \\

### 2010-04-18 doc/paper.tex
Add classifier results
git-svn-id: http://svn.scesoc.ca/evolution/trunk@91 fee98b4c-d1c3-4d34-9525-a468159e4a48
Added bibliography
git-svn-id: http://svn.scesoc.ca/evolution/trunk@92 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 
-\begin{table}
+\begin{table}[h]
 	\centering
@@
 
-\begin{figure}
+\begin{figure}[h]
 	\includegraphics[width=\textwidth]{fig-performance}
@@
 % ----------------------------------------------------------------------------
+The classifier was trained over 12000 generations with a final estimated lift of 4.51 (Figure~\ref{fig:evolution-lift}) out of a maximum lift of 5.
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.5\textwidth]{fig-evolution-lift}
+	\caption{Evolution of classifier lift. Note logarithmic generation number axis.}
+	\label{fig:evolution-lift}
+\end{figure}
 

### 2010-04-18 doc/paper.tex
Add some references
git-svn-id: http://svn.scesoc.ca/evolution/trunk@93 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 
-The CUDA programming model defines global and shared memory.
+The CUDA programming model defines global and shared memory \cite[\S3.2]{CudaProgGuide3.0}.
 Global memory is on-board but off-chip, expensive for random access and plentiful (100s of MB).
@@
 In addition, if groups of kernels access sequential addresses in global memory, the memory fetches are combined into a single transaction, maximizing bus bandwidth utilization.
-This is referred to as memory access coalescing.
+This is referred to as memory access coalescing \cite[\S5.3.2.1]{CudaProgGuide3.0}.
 
@@
 % ----------------------------------------------------------------------------
-Our demonstration classifier is designed for the 2009 UC San Diego Data Mining Contest "E-commerce Transaction Anomaly Classification" and with training data thereof.
+Our demonstration classifier is designed for the 2009 UC San Diego Data Mining Contest "E-commerce Transaction Anomaly Classification" \cite{UCSG-Contest} and with training data thereof.
 The training data consist of 94682 instances of 19 mixed features with a 1:50 binary class imbalance.

### 2010-04-18 doc/paper.tex
Add some more references
git-svn-id: http://svn.scesoc.ca/evolution/trunk@94 fee98b4c-d1c3-4d34-9525-a468159e4a48
Add Preliminary Data Analysis, move around performance results
git-svn-id: http://svn.scesoc.ca/evolution/trunk@95 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 % ----------------------------------------------------------------------------
+Fisher's linear discriminant \cite[\S 3.8.2]{Duda} obtains the 1D projection with best class separation by taking the eigenvector with the highest eigenvalue in a generalized eigenvector problem.
+We instead took the eigenvectors with the \emph{two} highest eigenvalues and plotted a 2D projection.
+This showed four defined clusters, which guided our choice of 4 hidden nodes with Gaussian radial basis functions.
 
@@
 As a result, meaningful numeric values can be extracted from arbitrary nominal values without domain knowledge relating to the values.
+This improves on the traditional ortogonal coding method \cite[slide 17]{MLP-NominalCoding} by avoiding allocation of a disproportionate number of input nodes to nominal features.
 
@@
 Of the operations presented in this paper, the MUTATE NODES and CROSSOVER WEIGHTS operations were determined to provide an optimal balance between performance and classifier accuracy, and have been implemented in our experiments.
-
-Preliminary analysis of test data has shown that a projection of the data onto a 2D space which maximize class separation produces four distinct eliptical regions.
-Therefore, test results have been obtained using an ANN which contains one layer of four radial basis function nodes.
-
-% Where does this fit?
-%  ANN outputs are treated as confidence values and used for ranking, and the lift of a classifier was directly used as the its fitness value for the GA.
 
@@
 Items 1, 2 and 3 were all implemented on the CUDA platform.
-Performance results are given as the combined time for one generation of these last three items but exclude one-time initialization.
+Performance results are given as the combined time for one generation of these three items but exclude one-time initialization.
 Item 4 is not very computationally intensive and was not parallelized.
@@
 This part of the process is particularly well-suited to the CUDA architecture because it has a very high ratio of mathematical operations to memory accesses (160 multiplications, 4 exponentiations and 80 bytes of memory accesses for our ANN topology).
+Parallelization is across individuals as well as training instances, which allows this part of the process to scale particularly well with hardware capabilities.
 
@@
 % ----------------------------------------------------------------------------
-To efficiently calculate the top 20\% threshold in a list of $n$ ANN output values, we can insert all values into a $k$-sized minheap where $k = 0.20 * n$, and take the root of the heap as the threshold.
+To efficiently calculate the top 20\% threshold in a list of $n$ unsorted ANN output values, we can insert all values into a $k$-sized binary minheap where $k = 0.20 * n$, and take the root of the heap as the threshold.
 Each minheap insertion costs $O(\log k)$ time.
@@
 % ----------------------------------------------------------------------------
-By keeping positive instances first in our training set, we avoid the need to consult the class of each instance (saving a costly global memory read).
-We simply iterate over the first $numPositives$ ANN outputs and count the outputs over the individual's 20\% threshold.
+By keeping positive instances first in our training set, we avoid the need to consult the class of each instance (global memory reads are costly).
+We simply iterate over the first $numPositives$ ANN outputs and count the outputs that are greater than the individual's 20\% threshold.
 Parallelization is across both individuals and groups of outputs.
-For each individual, we keep m counters which accumulate counts for indexes $i$ where $i \equiv 0 \bmod{m}$, $i \equiv 1 \bmod{m}$, ..., $i \equiv m - 1 \bmod{m}$.
+For each individual, we keep m counters which accumulate counts for output indices $i$ where $i \equiv 0 \bmod{m}$, $i \equiv 1 \bmod{m}$, ..., $i \equiv m - 1 \bmod{m}$.
 This enables parallelization even when there are not enough individuals in the population to fully occupy the device.
@@
 % ############################################################################
+We first present the computational performance of our algorithm for classifier training, followed by the estimated classification performance of the classifier designed for the UC San Diego data mining contest.
 
 % ----------------------------------------------------------------------------
-\subsection{Experimental Setup} \label{experiment}
+\subsection{Training Performance} \label{performance}
 % ----------------------------------------------------------------------------
@@
 
-% ----------------------------------------------------------------------------
-\subsection{Performance Results} \label{performance}
-% ----------------------------------------------------------------------------
 In most cases, the GPU implementation resulted in approximately an order of magnitude speedup over a CPU implementation (Figure~\ref{fig:training-performance}).
@@
 % ----------------------------------------------------------------------------
-\subsection{Classifier Results} \label{results}
+\subsection{Classification Performance} \label{results}
 % ----------------------------------------------------------------------------
@@
 By carefully designing memory organization, algorithm computational load and memory access patterns, we have obtained a 10-fold speedup compared to a conventional sequential CPU implementation.
-Additional workaround were necessary for the low amount of fast on-chip GPU memory.
+Additional workarounds were necessary for the low amount of fast on-chip GPU memory.
 Our method scales across population and training set sizes.

### 2010-04-18 doc/paper.tex
Rewrite preliminary data processing section, add reference to Duda
git-svn-id: http://svn.scesoc.ca/evolution/trunk@96 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 % ----------------------------------------------------------------------------
-Fisher's linear discriminant \cite[\S 3.8.2]{Duda} obtains the 1D projection with best class separation by taking the eigenvector with the highest eigenvalue in a generalized eigenvector problem.
-We instead took the eigenvectors with the \emph{two} highest eigenvalues and plotted a 2D projection.
-This showed four defined clusters, which guided our choice of 4 hidden nodes with Gaussian radial basis functions.
+The maximization of the ratio of between-class scatter to within-class scatter can be posed as a generalized eigenvalue problem where the eigenvalue with the largest corresponding eigenvalue is Fisher's linear discriminant \cite[\S 3.8.2]{Duda}.
+By taking instead the \emph{two} highest eigenvalues and plotting a 2D projection, we found that the data exhibited 4 defined clusters.
+This  guided our choice of 4 hidden nodes with Gaussian radial basis functions.
 

### 2010-04-18 doc/abstract.tex
Shortened version of paper.tex
git-svn-id: http://svn.scesoc.ca/evolution/trunk@97 fee98b4c-d1c3-4d34-9525-a468159e4a48

--- 2010-04-18 doc/paper.tex
@@
 The output is a single real value which can be interpreted as a classification confidence.
-ANN computation is dominated by the multiplications at the edges, proportional in number to the number of features (19) and the number of hidden nodes (4), and transcendental functions (exponentiation for Gaussian RBF) at the hidden nodes.
+ANN computation is dominated by the multiplications at the edges, proportional in number to the number of features and the number of hidden nodes, and transcendental functions (exponentiation for Gaussian RBF) at the hidden nodes.
 
@@
 NVidia's Compute Unified Device Architecture (CUDA) is a programming platform for massively parallel GPUs found in off-the-shelf graphics cards.
-GPUs consist of many independent floating-point units, providing significant speedup to data-parallel compute-intensive applications.
+GPUs consist of many independent floating-point units, providing significant speedup to data-parallel compute-intensive applications \cite{cuda}.
 Each unit is connected to on-board memory using a very wide bus, enabling high memory bandwidth provided certain particular memory access rules are respected.
 The combination of these features makes CUDA an ideal platform for GA-ANN training.
-
-The CUDA programming model defines global and shared memory \cite[\S3.2]{CudaProgGuide3.0}.
-Global memory is on-board but off-chip, expensive for random access and plentiful (100s of MB).
-Shared memory is on-chip, very fast for random access but very limited (16 KB).
-
-The computational unit in CUDA is the "kernel", which is a C function executed in parallel on the grid of available compute units in the CUDA device.
-The function is aware of its position in this grid such that it can apply the same algorithm to data at different addresses in memory.
-In addition, if groups of kernels access sequential addresses in global memory, the memory fetches are combined into a single transaction, maximizing bus bandwidth utilization.
-This is referred to as memory access coalescing \cite[\S5.3.2.1]{CudaProgGuide3.0}.
 
@@
 % ----------------------------------------------------------------------------
-Our demonstration classifier is designed for the 2009 UC San Diego Data Mining Contest "E-commerce Transaction Anomaly Classification" \cite{UCSG-Contest} and with training data thereof.
+Our demonstration classifier is designed for the 2009 UC San Diego Data Mining Contest "E-commerce Transaction Anomaly Classification" and with training data thereof.
 The training data consist of 94682 instances of 19 mixed features with a 1:50 binary class imbalance.
@@
 % ############################################################################
-
-% ----------------------------------------------------------------------------
-\subsection{Preliminary Data Analysis} \label{prelim}
-% ----------------------------------------------------------------------------
-
-% ----------------------------------------------------------------------------
-\subsection{Data Preprocessing Steps} \label{preprocessing}
-% ----------------------------------------------------------------------------
-We have implemented an algorithm which, given a tab seperated values file, converts nominal feature values in the file into numeric values.
-The algorithm replaces each nominal value in the file with the logarithm of the probability ratio between the positive and negative values observed given the nominal value.
-As a result, meaningful numeric values can be extracted from arbitrary nominal values without domain knowledge relating to the values.
-
-% ----------------------------------------------------------------------------
-\subsection{Classifier Design} \label{design}
-% ----------------------------------------------------------------------------
-In order to adapt to varying data sets, we have developped a flexible ANN model, allowing for an arbitrary number of hidden layers and nodes per layer (within the limits of GPU memory).
-All hidden layers must contain the same number of nodes, however input and output weights may be adjusted through the genetic algorithm so as to effectively reduce the number of nodes on a specific layer.
-
-The genetic mutation and crossover operations which we have implemented are based on the methods defined by Montana and Lawrence \cite{GA-ANN}.
-Of the operations presented in this paper, the MUTATE NODES and CROSSOVER WEIGHTS operations were determined to provide an optimal balance between performance and classifier accuracy, and have been implemented in our experiments.
-
-Preliminary analysis of test data has shown that a projection of the data onto a 2D space which maximize class separation produces four distinct eliptical regions.
-Therefore, test results have been obtained using an ANN which contains one layer of four radial basis function nodes.
-
-% Where does this fit?
-%  ANN outputs are treated as confidence values and used for ranking, and the lift of a classifier was directly used as the its fitness value for the GA.
 
@@
 
-% ----------------------------------------------------------------------------
-\subsection{Classifier Results} \label{results}
-% ----------------------------------------------------------------------------
-The classifier was trained over 12000 generations with a final estimated lift of 4.51 (Figure~\ref{fig:evolution-lift}) out of a maximum lift of 5.
-
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=0.5\textwidth]{fig-evolution-lift}
-	\caption{Evolution of classifier lift. Note logarithmic generation number axis.}
-	\label{fig:evolution-lift}
-\end{figure}
-
 % ############################################################################
@@
 Our method scales across population and training set sizes.
-
-% ############################################################################
-\section{Discussion} \label{disc}
-% ############################################################################
-
-% ----------------------------------------------------------------------------
-\subsection{Future Work} \label{future}
-% ----------------------------------------------------------------------------
-Montana and Lawrence \cite{GA-ANN} have demonstrated that a brief period of hill climbing after training with a genetic algorithm may increase the accuracy of classifier results.
-Thus, we could improve the accuracy of our methods by incorporating hill-climbing methods, such as the parallel backpropagation method described by Oei, Friedland and Janin \cite{backprop}.
 

### 2010-04-18 doc/abstract.tex
Fix small wording mistake
git-svn-id: http://svn.scesoc.ca/evolution/trunk@99 fee98b4c-d1c3-4d34-9525-a468159e4a48
Abstract good for review now?
git-svn-id: http://svn.scesoc.ca/evolution/trunk@100 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 ANNs are composed of a network of nodes connected by weighted directed edges.
-In a radial basis function (RBF) network architecture, nodes are organized in an input layer, a hidden layers, and an output layer.
+In a radial basis function (RBF) network architecture, nodes are organized in an input layer, hidden layers, and an output layer.
 Each layer is fully connected to the next.
-The nodes in the hidden layer calculate the distance between their inputs and a centre parameter, and pass a weighted distance to an RBF.
+The nodes in the hidden layer calculate the distance between their inputs and a center parameter, and pass a weighted distance to an RBF.
 The output node is a summer weighted by the incoming edges.
 The output is a single real value which can be interpreted as a classification confidence.
-ANN computation is dominated by the multiplications at the edges, proportional in number to the number of features and the number of hidden nodes, and transcendental functions (exponentiation for Gaussian RBF) at the hidden nodes.
+ANN computation is dominated by the multiplications at the edges, proportional to the number of features and the number of hidden nodes, and transcendental functions (exponentiation for Gaussian RBF) at the hidden nodes.
+ANN training has been implemented previously on GPU hardware, however these efforts have focused on training through backpropagation \cite{backprop}.
 
@@
 % ----------------------------------------------------------------------------
-Genetic algorithms (GAs) use multiple candidate solutions to simultaneously explore the search space of ANN parameters.
+Genetic algorithms (GAs) use multiple candidate solutions to simultaneously explore the search space of ANN parameters \cite{GA-ANN}.
 The "individuals" are randomly mutated, mated and selected during each of several generations.
@@
 Items 1, 2 and 3 were all implemented on the CUDA platform.
-Performance results are given as the combined time for one generation of these last three items but exclude one-time initialization.
+Performance results are given as the combined time for one generation of these first three items but exclude one-time initialization.
 Item 4 is not very computationally intensive and was not parallelized.
@@
 Feature data are organized to enable memory read coalescing.
-This part of the process is particularly well-suited to the CUDA architecture because it has a very high ratio of mathematical operations to memory accesses (160 multiplications, 4 exponentiations and 80 bytes of memory accesses for our ANN topology).
+This part of the process is particularly well-suited to the CUDA architecture because it has a very high ratio of mathematical operations to memory accesses (160 multiplications, 4 exponentiations and 80 bytes of memory accesses for the ANN topology used during testing).
 
@@
 % ############################################################################
-\bibliographystyle{plain}
+\bibliographystyle{unsrt}
 \bibliography{bibliography}     %loads bibliography.bib

### 2010-04-18 doc/paper.tex
Abstract good for review now?
git-svn-id: http://svn.scesoc.ca/evolution/trunk@100 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 
-The CUDA programming model defines global and shared memory \cite[\S3.2]{CudaProgGuide3.0}.
+The CUDA programming model defines global and shared memory.
 Global memory is on-board but off-chip, expensive for random access and plentiful (100s of MB).
@@
 In addition, if groups of kernels access sequential addresses in global memory, the memory fetches are combined into a single transaction, maximizing bus bandwidth utilization.
-This is referred to as memory access coalescing \cite[\S5.3.2.1]{CudaProgGuide3.0}.
+This is referred to as memory access coalescing.
 
@@
 
-The genetic mutation and crossover operations which we have implemented are based on the methods defined by Montana and Lawrence \cite{GA-ANN}.
-Of the operations presented in this paper, the MUTATE NODES and CROSSOVER WEIGHTS operations were determined to provide an optimal balance between performance and classifier accuracy, and have been implemented in our experiments.
+The classifier receives as inputs one numeric value for each feature of a pattern, each normalized to a value range between 0 and 1, and outputs a single numeric value which represents the confidence with which the pattern is a member of the positive class.
+These confidence values can be used for ranking during lift calculations, or a threshold for positive identification can be used to determine sensitivity or specificity measures.
 
@@
 
-% ############################################################################
-\section{Discussion} \label{disc}
-% ############################################################################
-
 % ----------------------------------------------------------------------------

### 2010-04-18 doc/paper.tex
Small wording fixes
git-svn-id: http://svn.scesoc.ca/evolution/trunk@101 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 % ----------------------------------------------------------------------------
-ANNs are composed of a network of nodes connected by weighted directed edges.
+ANNs are composed of nodes connected by weighted directed edges.
 In a radial basis function (RBF) network architecture, nodes are organized in an input layer, a hidden layers, and an output layer.
@@
 The nodes in the hidden layer calculate the distance between their inputs and a centre parameter, and pass a weighted distance to an RBF.
-The output node is a summer weighted by the incoming edges.
-The output is a single real value which can be interpreted as a classification confidence.
-ANN computation is dominated by the multiplications at the edges, proportional in number to the number of features (19) and the number of hidden nodes (4), and transcendental functions (exponentiation for Gaussian RBF) at the hidden nodes.
+The output node is a summer weighted by its incoming edges.
+The output is a single real value which is interpreted as a classification confidence.
+ANN computation is dominated by the multiplications at the edges, proportional in number to the number of features (19) and the number of hidden nodes (4), and the transcendental functions (exponentiation for Gaussian RBF) at the hidden nodes.
 
@@
 Each unit is connected to on-board memory using a very wide bus, enabling high memory bandwidth provided certain particular memory access rules are respected.
-The combination of these features makes CUDA an ideal platform for GA-ANN training.
+These features make CUDA an ideal platform for GA-ANN training.
 

### 2010-04-18 doc/abstract.tex
Small wording tweaks
git-svn-id: http://svn.scesoc.ca/evolution/trunk@102 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 % ----------------------------------------------------------------------------
-ANNs are composed of a network of nodes connected by weighted directed edges.
-In a radial basis function (RBF) network architecture, nodes are organized in an input layer, hidden layers, and an output layer.
+ANNs are composed of nodes connected by weighted directed edges.
+In a radial basis function (RBF) network architecture, nodes are organized in an input layer, a hidden layer, and an output layer.
 Each layer is fully connected to the next.
-The nodes in the hidden layer calculate the distance between their inputs and a center parameter, and pass a weighted distance to an RBF.
-The output node is a summer weighted by the incoming edges.
-The output is a single real value which can be interpreted as a classification confidence.
-ANN computation is dominated by the multiplications at the edges, proportional to the number of features and the number of hidden nodes, and transcendental functions (exponentiation for Gaussian RBF) at the hidden nodes.
+The nodes in the hidden layer calculate the distance between their inputs and a centre parameter, and pass a weighted distance to an RBF.
+The output node is a summer weighted by its incoming edges.
+The output is a single real value which is interpreted as a classification confidence.
+ANN computation is dominated by the number of multiplications at the edges, proportional to the number of features and the number of hidden nodes, and the transcendental functions (exponentiation for Gaussian RBF) at the hidden nodes.
 ANN training has been implemented previously on GPU hardware, however these efforts have focused on training through backpropagation \cite{backprop}.
@@
 Our demonstration classifier is designed for the 2009 UC San Diego Data Mining Contest "E-commerce Transaction Anomaly Classification" \cite{UCSG-Contest} and with training data thereof.
-The training data consist of 94682 instances of 19 mixed features with a 1:50 binary class imbalance.
+The training data consist of 94682 instances with 19 mixed features and a 1:50 binary class imbalance.
 The evaluation metric is "lift at 20\%", which can be understood as the ratio of the true positive rate in the top 20\% ranked instances to the overall positive rate of the data set.
@@
 Items 1, 2 and 3 were all implemented on the CUDA platform.
-Performance results are given as the combined time for one generation of these first three items but exclude one-time initialization.
+Performance results are given as the combined time for one generation of these three items but exclude one-time initialization.
 Item 4 is not very computationally intensive and was not parallelized.
@@
 This part of the process is particularly well-suited to the CUDA architecture because it has a very high ratio of mathematical operations to memory accesses (160 multiplications, 4 exponentiations and 80 bytes of memory accesses for the ANN topology used during testing).
+Parallelization is across individuals as well as training instances, which allows this part of the process to scale particularly well with hardware capabilities.
 

### 2010-04-18 doc/paper.tex
Small wording tweaks
git-svn-id: http://svn.scesoc.ca/evolution/trunk@102 fee98b4c-d1c3-4d34-9525-a468159e4a48

--- 2010-04-18 doc/paper.tex
@@
 % ----------------------------------------------------------------------------
-ANNs are composed of a network of nodes connected by weighted directed edges.
-In a radial basis function (RBF) network architecture, nodes are organized in an input layer, a hidden layers, and an output layer.
+ANNs are composed of nodes connected by weighted directed edges.
+In a radial basis function (RBF) network architecture, nodes are organized in an input layer, hidden layers, and an output layer.
 Each layer is fully connected to the next.
-The nodes in the hidden layer calculate the distance between their inputs and a centre parameter, and pass a weighted distance to an RBF.
+The nodes in the hidden layer calculate the distance between their inputs and a center parameter, and pass a weighted distance to an RBF.
 The output node is a summer weighted by the incoming edges.
 The output is a single real value which can be interpreted as a classification confidence.
-ANN computation is dominated by the multiplications at the edges, proportional in number to the number of features (19) and the number of hidden nodes (4), and transcendental functions (exponentiation for Gaussian RBF) at the hidden nodes.
+ANN computation is dominated by the multiplications at the edges, proportional to the number of features and the number of hidden nodes, and transcendental functions (exponentiation for Gaussian RBF) at the hidden nodes.
+ANN training has been implemented previously on GPU hardware, however these efforts have focused on training through backpropagation \cite{backprop}.
 
@@
 NVidia's Compute Unified Device Architecture (CUDA) is a programming platform for massively parallel GPUs found in off-the-shelf graphics cards.
-GPUs consist of many independent floating-point units, providing significant speedup to data-parallel compute-intensive applications.
+GPUs consist of many independent floating-point units, providing significant speedup to data-parallel compute-intensive applications \cite{cuda}.
 Each unit is connected to on-board memory using a very wide bus, enabling high memory bandwidth provided certain particular memory access rules are respected.
-The combination of these features makes CUDA an ideal platform for GA-ANN training.
+These features make CUDA an ideal platform for GA-ANN training.
 
@@
 Our demonstration classifier is designed for the 2009 UC San Diego Data Mining Contest "E-commerce Transaction Anomaly Classification" and with training data thereof.
-The training data consist of 94682 instances of 19 mixed features with a 1:50 binary class imbalance.
+The training data consist of 94682 instances with 19 mixed features and a 1:50 binary class imbalance.
 The evaluation metric is "lift at 20\%", which can be understood as the ratio of the true positive rate in the top 20\% ranked instances to the overall positive rate of the data set.
@@
 Feature data are organized to enable memory read coalescing.
-This part of the process is particularly well-suited to the CUDA architecture because it has a very high ratio of mathematical operations to memory accesses (160 multiplications, 4 exponentiations and 80 bytes of memory accesses for our ANN topology).
+This part of the process is particularly well-suited to the CUDA architecture because it has a very high ratio of mathematical operations to memory accesses (160 multiplications, 4 exponentiations and 80 bytes of memory accesses for the ANN topology used during testing).
 Parallelization is across individuals as well as training instances, which allows this part of the process to scale particularly well with hardware capabilities.

### 2010-04-19 doc/abstract.tex
Pages 1 and 2 of the abstract, now updated
git-svn-id: http://svn.scesoc.ca/evolution/trunk@103 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 
-\title{Accelerating Genetic ANN Training using the CUDA Platform}
+\title{CUDA-accelerated Genetic Feedforward-ANN Training for Data Mining}
 
@@
 By carefully mapping the problem onto the unique GPU architecture, we achieve order-of-magnitude speedup over a conventional CPU implementation.
-Furthermore, we show that the speedup is consistent across a wide range of population and training set sizes.
-This performance boost enables classifier design to span a wider range of population sizes and number of generations affording better classifiers and improved understanding of long-term trends in genetic algorithm convergence.
-Finally, we demonstrate this method in the context of the 2009 UC San Diego Data Mining Contest.
+Furthermore, we show that the speedup is consistent across a wide range of data set sizes, making this implementation ideal for large data sets.
+This performance boost enables the genetic algorithm to search a larger subset of the solution space, which results in a more accurate pattern classifier.
+Finally, we demonstrate this method in the context of the 2009 UC San Diego Data Mining Contest, achieving a world-class lift on a data set of 94682 e-commerce transactions.
 \end{abstract}
@@
 Genetic algorithms (GAs) are a stochastic, evolutionary approach to machine learning classifier training.
-While greedy methods can get stuck at local extrema, GAs  are theoretically capable of asymptotically reaching the global optimum.
-However, because they require the testing of a population of classifiers at each iteration, and the algorithm is repeated for many generations, they are extremely compute-intensive.
+While greedy methods can get stuck at local extrema, GAs  are theoretically capable of asymptotically reaching the global optimum \cite{GA-ANN}.
+However, because they require the evaluation of approximately 50 to 100 classifiers at each iteration, and the algorithm is repeated for several thousand generations, they are extremely compute-intensive.
 This is exacerbated by the use of artificial neural networks (ANNs) as the classifier because ANNs themselves are very compute-intensive.
@@
 ANNs are composed of nodes connected by weighted directed edges.
-In a radial basis function (RBF) network architecture, nodes are organized in an input layer, a hidden layer, and an output layer.
+In a feed-forward network architecture, nodes are organized in an input layer, several hidden layers, and an output layer.
 Each layer is fully connected to the next.
-The nodes in the hidden layer calculate the distance between their inputs and a centre parameter, and pass a weighted distance to an RBF.
-The output node is a summer weighted by its incoming edges.
+Each of the hidden nodes at hidden layer $n$ perform a transcendental function with inputs equal to the outputs of layer $n-1$ each multiplied by the weight on the edge through which they are fed.
+In a radial basis function (RBF) network, the nodes in the hidden layer calculate the distance between their inputs and a centre vector, and pass a weighted distance to a RBF.
+Nodes in a gaussian function network perform a similar operation with a gaussian function in place of the RBF.
+The output node is typically linear, computing a weighted sum of its inputs.
 The output is a single real value which is interpreted as a classification confidence.
-ANN computation is dominated by the number of multiplications at the edges, proportional to the number of features and the number of hidden nodes, and the transcendental functions (exponentiation for Gaussian RBF) at the hidden nodes.
-ANN training has been implemented previously on GPU hardware, however these efforts have focused on training through backpropagation \cite{backprop}.
+ANN computation is dominated by the number of multiplications at the edges between nodes, proportional to the number of features and the number of hidden nodes, and by the transcendental functions (exponentiation for RBF) at the hidden nodes.
+ANN training has been implemented previously on GPU hardware, however these efforts have focused on training through backpropagation \cite{backprop} as opposed to genetic algorithms.
 
@@
 The "individuals" are randomly mutated, mated and selected during each of several generations.
-Mutation and mating is done using a problem-specific representation of candidate solutions and "genetic operators." Selection is performed by applying the classifier to the training set and calculating a fitness value (classifier performance).
+Mutation and mating is done using a problem-specific representation of candidate solutions and "genetic operators." Selection is performed by applying the classifier to the training set and calculating a fitness value (classifier accuracy).
 The competitive bias imposed by selection attempts to mimic the "survival of the fittest" principle so often seen in Nature.
+Because GAs blindly search a solution space and ANNs can be of arbitrary complexity, this technique is suitable for data mining applications using large or complex data sets.
 
-The computation required for GA training of ANNs is proportional to the number of generations ($10^1$-$10^4$), to the size of each generation ($10^1$-$10^2$ individuals) and to the training set size ($10^2$-$10^6$ instances).
+The computation required for GA training of ANNs is proportional to the number of generations ($10^1$-$10^4$), to the size of each generation ($10^1$-$10^2$ individuals) and to the size of the data set ($10^2$-$10^6$ instances).
 
@@
 NVidia's Compute Unified Device Architecture (CUDA) is a programming platform for massively parallel GPUs found in off-the-shelf graphics cards.
-GPUs consist of many independent floating-point units, providing significant speedup to data-parallel compute-intensive applications \cite{cuda}.
-Each unit is connected to on-board memory using a very wide bus, enabling high memory bandwidth provided certain particular memory access rules are respected.
+GPUs consist of several dozen independent floating-point units, providing significant speedup to data-parallel compute-intensive applications \cite{cuda}.
+Each unit is connected to on-board memory using a very wide bus, enabling high memory bandwidth provided certain memory access rules are respected.
 These features make CUDA an ideal platform for GA-ANN training.
@@
 % ----------------------------------------------------------------------------
-\subsection{Data Mining Contest} \label{contest}
+\subsection{Sample Application} \label{contest}
 % ----------------------------------------------------------------------------
 Our demonstration classifier is designed for the 2009 UC San Diego Data Mining Contest "E-commerce Transaction Anomaly Classification" \cite{UCSG-Contest} and with training data thereof.
-The training data consist of 94682 instances with 19 mixed features and a 1:50 binary class imbalance.
+The training data consist of 94682 instances with 19 features of mixed types and a 1:50 binary class imbalance.
 The evaluation metric is "lift at 20\%", which can be understood as the ratio of the true positive rate in the top 20\% ranked instances to the overall positive rate of the data set.
+Lift at 20\% is commonly used in marketing research in order to select the subset of a population which is most likely to respond to solicitation.
 
@@
 % ----------------------------------------------------------------------------
-\subsection{Classifier Implementation} \label{implementation}
+\subsection{Implementation of GA Training of ANNs} \label{implementation}
 % ----------------------------------------------------------------------------
-Each GA iteration consists of the following steps:
+Each iteration of the GA (referred to as a generation) consists of the following steps:
 \begin{enumerate}
-	\item Computation of ANN outputs
-	\item Calculation of top 20\% threshold value (each individual)
+	\item Computation of ANN outputs, $i.e.$ classification confidence for each instance in the data set as calculated by each candidate ANN in the generation
+	\item Calculation of top 20\% threshold value for each candidate ANN
 	\item Counting of the positive instances with outputs above this threshold (each individual)

### 2010-04-19 doc/abstract.tex
More updating of page 1 and 2
git-svn-id: http://svn.scesoc.ca/evolution/trunk@104 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 % ----------------------------------------------------------------------------
-\subsection{Genetic Algorithms} \label{ga}
+\subsection{Genetic Algorithms as ANN Training Agents} \label{ga}
 % ----------------------------------------------------------------------------
 Genetic algorithms (GAs) use multiple candidate solutions to simultaneously explore the search space of ANN parameters \cite{GA-ANN}.
-The "individuals" are randomly mutated, mated and selected during each of several generations.
-Mutation and mating is done using a problem-specific representation of candidate solutions and "genetic operators." Selection is performed by applying the classifier to the training set and calculating a fitness value (classifier accuracy).
+The candidate ANNs are randomly mutated, mated and selected during each of several generations.
+Mutation and mating is done using a problem-specific representation of candidate solutions and "genetic operators." Selection requires calculating the fitness of each candidate ANN (a classifier accuracy metric such as sensitivity) then preferentially selecting candidate ANNs based on fitness.
 The competitive bias imposed by selection attempts to mimic the "survival of the fittest" principle so often seen in Nature.

### 2010-04-19 doc/abstract.tex
Found another mistake
git-svn-id: http://svn.scesoc.ca/evolution/trunk@105 fee98b4c-d1c3-4d34-9525-a468159e4a48
Address JRG's comments to section Methods and following
git-svn-id: http://svn.scesoc.ca/evolution/trunk@106 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 \begin{abstract}
-We present an implementation of genetic algorithm (GA) training of artificial neural networks (ANNs) targeting commodity graphics cards (GPUs).
+We present an implementation of genetic algorithm (GA) training of feedforward artificial neural networks (ANNs) targeting commodity graphics cards (GPUs).
 By carefully mapping the problem onto the unique GPU architecture, we achieve order-of-magnitude speedup over a conventional CPU implementation.
@@
 In a radial basis function (RBF) network, the nodes in the hidden layer calculate the distance between their inputs and a centre vector, and pass a weighted distance to a RBF.
-Nodes in a gaussian function network perform a similar operation with a gaussian function in place of the RBF.
+Nodes in a sigmoid function network perform a similar operation with a sigmoid function in place of the RBF.
 The output node is typically linear, computing a weighted sum of its inputs.
@@
 % ############################################################################
-\section{Methods} \label{algimp}
+\section{Implementation of GA Training of ANNs} \label{algimp}
 % ############################################################################
 
-% ----------------------------------------------------------------------------
-\subsection{Implementation of GA Training of ANNs} \label{implementation}
-% ----------------------------------------------------------------------------
-Each iteration of the GA (referred to as a generation) consists of the following steps:
+Our GA begins by generating an initial population of random individuals, each of which is a set parameters describing one ANN.
+Feature values for all training instances are loaded into GPU global memory in preparation for ANN computations.
+Then, the following steps are executed in sequence for each generation of the GA:
+
 \begin{enumerate}
-	\item Computation of ANN outputs, $i.e.$ classification confidence for each instance in the data set as calculated by each candidate ANN in the generation
-	\item Calculation of top 20\% threshold value for each candidate ANN
-	\item Counting of the positive instances with outputs above this threshold (each individual)
-	\item Genetic algorithm processing -- mating, mutation, selection
+	\item Compute the ANN output corresponding to each training instance, as calculated by each candidate ANN in the population.
+	\item Find the threshold that separates the highest 20\% outputs for each candidate's set of output values.
+	\item Compute the number of top 20\% instances which are truly positive, again for each candidate ANN. This allows us to calculate lift, which is used as the fitness value for each candidate ANN.
+	\item Apply genetic operators: mating, mutation and selection. The selected candidate ANNs become the new population, to be used in the next generation.
 \end{enumerate}
 
-Items 1, 2 and 3 were all implemented on the CUDA platform.
+Items 1, 2 and 3 were all implemented on the CUDA platform and are described in detail below.
 Performance results are given as the combined time for one generation of these three items but exclude one-time initialization.
@@
 % ----------------------------------------------------------------------------
-ANN output values are computed by reading feature values and applying the appropriate operations represented by the network topology and parameters.
-Feature data are organized to enable memory read coalescing.
-This part of the process is particularly well-suited to the CUDA architecture because it has a very high ratio of mathematical operations to memory accesses (160 multiplications, 4 exponentiations and 80 bytes of memory accesses for the ANN topology used during testing).
-Parallelization is across individuals as well as training instances, which allows this part of the process to scale particularly well with hardware capabilities.
+ANN output values are computed by reading the training instances as real-valued feature vectors and applying the appropriate operations represented by the network topology and parameters.
+Feature data are organized to efficiently use memory bandwidth.
+ANN output calculation is particularly well-suited to the CUDA architecture because it has a very high ratio of mathematical operations to memory accesses (160 multiplications, 4 exponentiations and 80 bytes of memory accesses for our ANN topology).
+In addition, each output value depends only on the current ANN parameters and one training instance's feature values.
+Therefore, parallelization of this step scales particularly well with hardware capabilities.
 
@@
 % ----------------------------------------------------------------------------
-To efficiently calculate the top 20\% threshold in a list of $n$ unsorted ANN output values, we can insert all values into a $k$-sized binary minheap where $k = 0.20 * n$, and take the root of the heap as the threshold.
-Each minheap insertion costs $O(\log k)$ time.
-Because device shared memory (16~KB) is not large enough for our value of $k$, we instead use $p$ passes of $\frac{k}{p}$-sized heaps with which we find the top 5\% threshold, then the top 10\%, etc.
+Next, we must calculate the $k^{th}$ largest value in each candidate's list of output values, where $k = 0.20 * n$ and $n$ is the number of output values (number of training instances).
+We can accomplish this by iterating through all $n$ unsorted output values and inserting each into a binary minheap of fixed size $k$.
+The heap size is kept fixed by removing the root (smallest node) after each insertion.
+At the end of this process, the root of the heap is the $k^{th}$ largest output value.
+The insertion and removal can be combined into a single operation which costs $O(\log k)$ time.
+The heap occupies $4 * k$ bytes of device shared memory.
+
+Because fast on-chip memory (16~KB) is not large enough for our value of $k$ ($4 * 18936 =$ 74~KB), we instead use $p$ passes of $\frac{k}{p}$-sized heaps.
+Each pass examines only values which are less than the threshold calculated by the preceeding pass.
+For example, the third pass examines only values below the 10\% threshold and finds the top 5\% threshold of this subset of values.
+This threshold however is the 15\% threshold of the full list.
+The threshold calculated by the last pass is the desired 20\% threshold.
+
 The overall complexity of our algorithm is $O(np \log \frac{k}{p})$.
-Parallelization is across individuals.
+Each top 20\% calculation is performed independently for each candidate ANN.
+Therefore, the parallel performance of this step scales with hardware capabilities when there are enough candidates to occupy the entire device.
 
 % ----------------------------------------------------------------------------
-\subsection{Counting of Positives Above Threshold} \label{implementation}
+\subsection{Computation of Number of Truly Positive Top 20\% Instances} \label{implementation}
 % ----------------------------------------------------------------------------
-By keeping positive instances first in our training set, we avoid the need to consult the class of each instance (global memory reads are costly).
-We simply iterate over the first $numPositives$ ANN outputs and count the outputs that are greater than the individual's 20\% threshold.
-Parallelization is across both individuals and groups of outputs.
-For each individual, we keep m counters which accumulate counts for output indices $i$ where $i \equiv 0 \bmod{m}$, $i \equiv 1 \bmod{m}$, ..., $i \equiv m - 1 \bmod{m}$.
-This enables parallelization even when there are not enough individuals in the population to fully occupy the device.
-Each batch of $m$ kernels accesses $m$ sequential locations in memory, which results in coalesced reads.
+To compute the number of truly positive instances in the top 20\% of scores, if training instances are unsorted, we must read from memory each instance's true class and corresponding ANN output value and compare the output value to the candidate's threshold.
+For $n$ training instances, this requires $2 * n$ memory accesses.
+However, by keeping the $n_p$ positive instances first in memory, we need only read the first $n_p$ ANN outputs, knowing that they correspond to the positive instances.
+This reduces the number of memory accesses to $n_p$, which is significant particularly for training data with low positive rate (in our case, $2 * n = 189364$ and $n_p = 2080$).
+Sorting the training instances by class is a cheap operation and can be done by the host CPU as part of initialization.
+
+Counting of positive instances above the candidate's threshold is performed independently, in parallel, for each candidate.
+However, because there may not be enough individuals to fully occupy the GPU, we also split the counting process for each individual into 256 independent sub-counts: the first sub-counter processes training instances 0, 256, 512, \ldots, the second sub-counter instances 1, 257, 513, \ldots, such that the sub-counts can be executed in parallel.
+The striped assignment of training instance indices to sub-counters is necessary to take advantage of the full memory bandwidth of the GPU.
+The host CPU then aggregates the sub-counts by summing, which is a very lightweight operation.
+
+Finally, each count is divided by the number of positive instances to give a positive rate in the top 20\%.
+Then, the ratio of this positive rate to the overall positive rate is the lift of each individual.
+These operations are also comparatively lightweight and executed on the host CPU.
 
@@
 	\centering
-	\begin{tabular}{ll}
-	\textbf{Legend Label} & \textbf{Hardware} \\
+	\begin{tabular}{lllll}
+	\textbf{Legend Label} & \textbf{Hardware} & \textbf{Clock} & \textbf{Free RAM} & \textbf{OS} \\
 	\hline
-	Core2 & Intel Core 2 Q9450 CPU \\
-	i7 & Intel i7 920 CPU \\
-	GTX260 & NVidia GTX260 GPU \\
-	GTX275 & NVidia GTX275 GPU \\
+	Core2  & Intel Core 2 Q9450 CPU & 2.66~GHz & 7.5~GB & Linux 2.6 (64-bit) \\
+	i7     & Intel i7 920 CPU       & ?        & ?      & Windows 7 (64-bit) \\
+	GTX260 & NVidia GTX260 GPU      & ?        & N/A    & Windows 7 (64-bit) \\
+	GTX275 & NVidia GTX275 GPU      & 1.4~GHz  & N/A    & Linux 2.6 (64-bit) \\
 	\hline
 	\end{tabular}
-	\caption{Experimental hardware}
+	\caption{Experimental hardware. Free RAM is OS-reported available memory before running each experiment.}
 	\label{tab:experimental-hardware}
@@
 In most cases, the GPU implementation resulted in approximately an order of magnitude speedup over a CPU implementation (Figure~\ref{fig:training-performance}).
-For low population sizes, GPU initialization and data copying dominates.
+For population sizes below 10, observed speedup decreases and no speedup is observed in the single-individual case.
+This is due to GPU initialization and data copying dominating execution time.
+However, in these cases, both implementation are still quite fast (20~ms per generation).
 
 \begin{figure}[h]
+	\centering
 	\includegraphics[width=\textwidth]{fig-performance}
@@
 By carefully designing memory organization, algorithm computational load and memory access patterns, we have obtained a 10-fold speedup compared to a conventional sequential CPU implementation.
-Additional workarounds were necessary for the low amount of fast on-chip GPU memory.
-Our method scales across population and training set sizes.
+A multipass approach was required for part of our system to compensate for the limited amount of fast on-chip GPU memory.
+Our method scales across population and training set sizes and is expected to be useful in other data intensive machine learning and data mining tasks.
 

### 2010-04-19 doc/abstract.tex
A few individual->candidate and another small typo
git-svn-id: http://svn.scesoc.ca/evolution/trunk@107 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 
-The computation required for GA training of ANNs is proportional to the number of generations ($10^1$-$10^4$), to the size of each generation ($10^1$-$10^2$ individuals) and to the size of the data set ($10^2$-$10^6$ instances).
+The computation required for GA training of ANNs is proportional to the number of generations ($10^1$-$10^4$), to the size of each generation ($10^1$-$10^2$ candidates) and to the size of the data set ($10^2$-$10^6$ instances).
 
@@
 
-Our GA begins by generating an initial population of random individuals, each of which is a set parameters describing one ANN.
+Our GA begins by generating an initial population of random candidate ANNs, each of which is a set parameters describing one ANN.
 Feature values for all training instances are loaded into GPU global memory in preparation for ANN computations.
@@
 Counting of positive instances above the candidate's threshold is performed independently, in parallel, for each candidate.
-However, because there may not be enough individuals to fully occupy the GPU, we also split the counting process for each individual into 256 independent sub-counts: the first sub-counter processes training instances 0, 256, 512, \ldots, the second sub-counter instances 1, 257, 513, \ldots, such that the sub-counts can be executed in parallel.
+However, because there may not be enough candidates to fully occupy the GPU, we also split the counting process for each candidate into 256 independent sub-counts: the first sub-counter processes training instances 0, 256, 512, \ldots, the second sub-counter instances 1, 257, 513, \ldots, such that the sub-counts can be executed in parallel.
 The striped assignment of training instance indices to sub-counters is necessary to take advantage of the full memory bandwidth of the GPU.
@@
 Finally, each count is divided by the number of positive instances to give a positive rate in the top 20\%.
-Then, the ratio of this positive rate to the overall positive rate is the lift of each individual.
+Then, the ratio of this positive rate to the overall positive rate is the lift of each candidate.
 These operations are also comparatively lightweight and executed on the host CPU.
@@
 In most cases, the GPU implementation resulted in approximately an order of magnitude speedup over a CPU implementation (Figure~\ref{fig:training-performance}).
-For population sizes below 10, observed speedup decreases and no speedup is observed in the single-individual case.
+For population sizes below 10, observed speedup decreases and no speedup is observed in the single-candidate case.
 This is due to GPU initialization and data copying dominating execution time.
-However, in these cases, both implementation are still quite fast (20~ms per generation).
+However, in these cases, both implementations are still quite fast (20~ms per generation).
 
@@
 % ############################################################################
-We have presented an implementation of GA training of ANN classifiers for the CUDA platform for GPU programming.
+We have presented an implementation of GA training of feedforward ANN classifiers for the CUDA platform for GPU programming.
 By carefully designing memory organization, algorithm computational load and memory access patterns, we have obtained a 10-fold speedup compared to a conventional sequential CPU implementation.

### 2010-04-19 doc/abstract.tex
Sending this version of the abstract off to Green
git-svn-id: http://svn.scesoc.ca/evolution/trunk@108 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
+\textit{}
 % ===========================================================================
@@
 Furthermore, we show that the speedup is consistent across a wide range of data set sizes, making this implementation ideal for large data sets.
-This performance boost enables the genetic algorithm to search a larger subset of the solution space, which results in a more accurate pattern classifier.
+This performance boost enables the genetic algorithm to search a larger subset of the solution space, which results in more accurate pattern classification.
 Finally, we demonstrate this method in the context of the 2009 UC San Diego Data Mining Contest, achieving a world-class lift on a data set of 94682 e-commerce transactions.
@@
 Genetic algorithms (GAs) are a stochastic, evolutionary approach to machine learning classifier training.
-While greedy methods can get stuck at local extrema, GAs  are theoretically capable of asymptotically reaching the global optimum \cite{GA-ANN}.
-However, because they require the evaluation of approximately 50 to 100 classifiers at each iteration, and the algorithm is repeated for several thousand generations, they are extremely compute-intensive.
+While greedy methods will get stuck at local extrema, GAs  are theoretically capable of asymptotically reaching the global optimum \cite{GA-ANN}.
+However, because they require the evaluation of approximately 50 to 100 classifiers at each iteration, and the algorithm is repeated for several thousand iterations, they are extremely compute-intensive.
 This is exacerbated by the use of artificial neural networks (ANNs) as the classifier because ANNs themselves are very compute-intensive.
-Compared to a traditional sequential implementation of GA training of ANNs, the use of graphical processing units (GPUs) as parallel processors provides significant performance improvements, at a fraction of the price of alternatives such as cluster computing.
+
+Compared to a traditional sequential implementation of GA training of ANNs, the use of graphical processing units (GPUs) as parallel processors provides significant performance improvements at a fraction of the price of alternatives such as cluster computing.
 GPUs, however, use a specialized programming paradigm which must be taken into account to leverage their full processing power.
@@
 ANNs are composed of nodes connected by weighted directed edges.
-In a feed-forward network architecture, nodes are organized in an input layer, several hidden layers, and an output layer.
+In a feed-forward network architecture, nodes are organized into an input layer, several hidden layers, and an output layer.
 Each layer is fully connected to the next.
-Each of the hidden nodes at hidden layer $n$ perform a transcendental function with inputs equal to the outputs of layer $n-1$ each multiplied by the weight on the edge through which they are fed.
-In a radial basis function (RBF) network, the nodes in the hidden layer calculate the distance between their inputs and a centre vector, and pass a weighted distance to a RBF.
+Each of the hidden nodes at hidden layer $n$ perform a transcendental function with inputs equal to the outputs of layer $n-1$ each multiplied by the weight on the edge through which they are fed from layer $n-1$ to layer $n$.
+In a radial basis function (RBF) network, the nodes in each hidden layer calculate the distance between their inputs and a centre vector, and pass a weighted distance to a RBF.
 Nodes in a sigmoid function network perform a similar operation with a sigmoid function in place of the RBF.
 The output node is typically linear, computing a weighted sum of its inputs.
-The output is a single real value which is interpreted as a classification confidence.
-ANN computation is dominated by the number of multiplications at the edges between nodes, proportional to the number of features and the number of hidden nodes, and by the transcendental functions (exponentiation for RBF) at the hidden nodes.
+The output of the ANN is a single real value which is interpreted as a classification confidence.
+ANN computation is dominated by the number of multiplications at the edges between nodes, proportional to the number of features in the data set and the number of hidden nodes, and by the transcendental functions at the hidden nodes, which are typically exponentiations.
 ANN training has been implemented previously on GPU hardware, however these efforts have focused on training through backpropagation \cite{backprop} as opposed to genetic algorithms.
@@
 % ----------------------------------------------------------------------------
-Genetic algorithms (GAs) use multiple candidate solutions to simultaneously explore the search space of ANN parameters \cite{GA-ANN}.
+Genetic algorithms (GAs) use batches of candidate solutions to simultaneously explore the search space of ANN parameters \cite{GA-ANN}.
 The candidate ANNs are randomly mutated, mated and selected during each of several generations.
 Mutation and mating is done using a problem-specific representation of candidate solutions and "genetic operators." Selection requires calculating the fitness of each candidate ANN (a classifier accuracy metric such as sensitivity) then preferentially selecting candidate ANNs based on fitness.
-The competitive bias imposed by selection attempts to mimic the "survival of the fittest" principle so often seen in Nature.
+The competitive bias imposed by selection attempts to mimic the "survival of the fittest" principle often seen in nature.
 Because GAs blindly search a solution space and ANNs can be of arbitrary complexity, this technique is suitable for data mining applications using large or complex data sets.
@@
 
-Our GA begins by generating an initial population of random candidate ANNs, each of which is a set parameters describing one ANN.
+Our GA begins by generating an initial population of random candidate ANNs, each of which is a set of parameters describing one ANN.
 Feature values for all training instances are loaded into GPU global memory in preparation for ANN computations.
@@
 % ----------------------------------------------------------------------------
-\subsection{Computation of ANN Output Values} \label{implementation}
+\subsection{Computation of ANN Output Values} \label{ANNcomp}
 % ----------------------------------------------------------------------------
@@
 % ----------------------------------------------------------------------------
-\subsection{Calculation of Top 20\% Threshold} \label{implementation}
+\subsection{Calculation of Top 20\% Threshold} \label{top20}
 % ----------------------------------------------------------------------------
@@
 For example, the third pass examines only values below the 10\% threshold and finds the top 5\% threshold of this subset of values.
-This threshold however is the 15\% threshold of the full list.
+This threshold therefore is the 15\% threshold of the full list.
 The threshold calculated by the last pass is the desired 20\% threshold.
@@
 % ----------------------------------------------------------------------------
-\subsection{Computation of Number of Truly Positive Top 20\% Instances} \label{implementation}
+\subsection{Computation of Number of Truly Positive Top 20\% Instances} \label{truepos}
 % ----------------------------------------------------------------------------
@@
 	Core2  & Intel Core 2 Q9450 CPU & 2.66~GHz & 7.5~GB & Linux 2.6 (64-bit) \\
-	i7     & Intel i7 920 CPU       & ?        & ?      & Windows 7 (64-bit) \\
-	GTX260 & NVidia GTX260 GPU      & ?        & N/A    & Windows 7 (64-bit) \\
+	i7     & Intel i7 920 CPU       & 2.66~GHz & 2.4~GB & Windows 7 (64-bit) \\
+	GTX260 & NVidia GTX260 GPU      & 1.24~GHz & N/A    & Windows 7 (64-bit) \\
 	GTX275 & NVidia GTX275 GPU      & 1.4~GHz  & N/A    & Linux 2.6 (64-bit) \\

### 2010-04-20 doc/abstract.tex
Final version of the abstract!
git-svn-id: http://svn.scesoc.ca/evolution/trunk@109 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
-\textit{}
 % ===========================================================================
@@
 % ############################################################################
-Genetic algorithms (GAs) are a stochastic, evolutionary approach to machine learning classifier training.
+Genetic algorithms (GAs) are a stochastic, evolutionary approach to machine learning for pattern classification.
 While greedy methods will get stuck at local extrema, GAs  are theoretically capable of asymptotically reaching the global optimum \cite{GA-ANN}.
@@
 ANNs are composed of nodes connected by weighted directed edges.
-In a feed-forward network architecture, nodes are organized into an input layer, several hidden layers, and an output layer.
+In a feed-forward network architecture, nodes are organized into an input layer, one or more hidden layers, and an output layer.
 Each layer is fully connected to the next.
-Each of the hidden nodes at hidden layer $n$ perform a transcendental function with inputs equal to the outputs of layer $n-1$ each multiplied by the weight on the edge through which they are fed from layer $n-1$ to layer $n$.
+Each of the hidden nodes at hidden layer $n$ perform a transcendental function with inputs equal to the outputs of layer $n-1$ each multiplied by the weight associated with the edge through which they are fed from layer $n-1$ to layer $n$.
 In a radial basis function (RBF) network, the nodes in each hidden layer calculate the distance between their inputs and a centre vector, and pass a weighted distance to a RBF.
@@
 The output of the ANN is a single real value which is interpreted as a classification confidence.
-ANN computation is dominated by the number of multiplications at the edges between nodes, proportional to the number of features in the data set and the number of hidden nodes, and by the transcendental functions at the hidden nodes, which are typically exponentiations.
+ANN computation is dominated by the number of multiplications at the directed edges, proportional to the number of features in the data set and the number of hidden nodes, and by the transcendental functions at the hidden nodes, which are typically exponentiations.
 ANN training has been implemented previously on GPU hardware, however these efforts have focused on training through backpropagation \cite{backprop} as opposed to genetic algorithms.
@@
 % ----------------------------------------------------------------------------
-Genetic algorithms (GAs) use batches of candidate solutions to simultaneously explore the search space of ANN parameters \cite{GA-ANN}.
+Genetic algorithms (GAs) use populations of candidate solutions to simultaneously explore the search space of ANN parameters \cite{GA-ANN}.
 The candidate ANNs are randomly mutated, mated and selected during each of several generations.
-Mutation and mating is done using a problem-specific representation of candidate solutions and "genetic operators." Selection requires calculating the fitness of each candidate ANN (a classifier accuracy metric such as sensitivity) then preferentially selecting candidate ANNs based on fitness.
+Mutation and mating is done using a problem-specific representation of candidate solutions and "genetic operators." Selection requires calculating the fitness of each candidate ANN  then preferentially selecting candidate ANNs based on fitness.
+The fitness of a classifier is a classifier accuracy metric such as sensitivity.
 The competitive bias imposed by selection attempts to mimic the "survival of the fittest" principle often seen in nature.
-Because GAs blindly search a solution space and ANNs can be of arbitrary complexity, this technique is suitable for data mining applications using large or complex data sets.
+Because GAs perform an unbiased search of a solution space and ANNs can be of arbitrary complexity, this technique is suitable for data mining applications using large or complex data sets.
 
@@
 	\item Compute the ANN output corresponding to each training instance, as calculated by each candidate ANN in the population.
-	\item Find the threshold that separates the highest 20\% outputs for each candidate's set of output values.
+	\item Find the threshold that defines the highest 20\% outputs for each candidate's set of output values.
 	\item Compute the number of top 20\% instances which are truly positive, again for each candidate ANN. This allows us to calculate lift, which is used as the fitness value for each candidate ANN.
@@
 However, by keeping the $n_p$ positive instances first in memory, we need only read the first $n_p$ ANN outputs, knowing that they correspond to the positive instances.
-This reduces the number of memory accesses to $n_p$, which is significant particularly for training data with low positive rate (in our case, $2 * n = 189364$ and $n_p = 2080$).
-Sorting the training instances by class is a cheap operation and can be done by the host CPU as part of initialization.
+This reduces the number of memory accesses to $n_p$, which is significant particularly for training data with low positive rate (in our case, $2 * n = 189364$ while $n_p = 2080$).
+Sorting the training instances by class is a cheap operation which is performed only once during initialization and can be done by the host CPU.
 

### 2010-04-20 doc/paper.tex
Creating a seperate file for report.  Also, added a lot to the file.
git-svn-id: http://svn.scesoc.ca/evolution/trunk@110 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 % ----------------------------------------------------------------------------
-The classifier was trained over 12000 generations with a final estimated lift of 4.51 (Figure~\ref{fig:evolution-lift}) out of a maximum lift of 5.
+The classifier was trained over 12000 generations; the fitness metric during this training was lift at 20\% calculated on a training set which consisted of 70\% of the data set, chosen at random during initialization.
+After each generation, the candidate ANN with the highest fitness value was used to classify 100 random subsets of the hold-out data set, and the average of the lift at 20\% of these classifications was recorded.
+Figure~\ref{fig:evolution-lift} demonstrates the experimental results across 12,000 generations of population 50.
+The final estimated lift is 4.51 (Figure~\ref{fig:evolution-lift}) out of a maximum lift of 5.
 
@@
 \end{figure}
+
+When applied to the 2009 UC San Diego Data Mining Contest blind test data set, the resulting classifier produced a lift at 20\% value of 4.031, placing 35th in the world for classification of this data set.
+Due to the stochastic nature of genetic algorithms and the contest rules which state that contestants may test any number of candidate classifiers, it is likely that repeated entry into this contest would result in a higher ranking.
 

### 2010-04-20 doc/report.tex
Creating a seperate file for report.  Also, added a lot to the file.
git-svn-id: http://svn.scesoc.ca/evolution/trunk@110 fee98b4c-d1c3-4d34-9525-a468159e4a48

--- 2010-04-20 doc/abstract.tex
@@
 \evensidemargin=0.25in \topmargin=-0.5in
+\linespread{2}
 %--------------- Algorithm --------------------------------------------------
@@
 % ############################################################################
+\subsection{Classifier Design} \label{design}
+% ############################################################################
+
+While our GA-ANN classifier was designed for the 2009 UC San Diego Data Mining Contest data set, it is capable of adapting to a variety of data sets.
+We have developped a flexible ANN topology, allowing for one or two hidden layers and an arbitrary number of nodes per hidden layer (within the limits of GPU memory).
+In addition, each layer of the ANN is capable of performing RBF or sigmoid calculations.
+
+The classifier receives as inputs one numeric value for each feature in a data set, all of which are normalized to a value range between 0 and 1.
+The classifier outputs a single numeric value, also in the range of 0 and 1, which represents the confidence with which the pattern is a member of the positive class.
+These confidence values can be used for ranking during lift calculations, or for sensitivity or specificity measures with an arbitrary value chosen as a threshold for positive identification of a pattern.
+Therefore, the fitness function is not limited to lift at 20\% as presented in this paper.
+
+In our final design, each candidate ANN is represented by a chromosome which consists of four arrays of real-valued numbers.
+Edge weights between input nodes and hidden layers, edge weights between hidden and output layers, width values for RBFs at hidden nodes and center values for RBFs at hidden nodes are all encoded, and can take on any value.
+Figure~\ref{fig:evolution-lift} demonstrates the chromosome structure and its relation to ANN structure.
+This structure is simple yet powerful; individual edges are 'removed' by the genetic algorithm when their weight is set to zero and nodes are 'removed' by the genetic algorithm when their RBF width value is set to zero.
+Thus, we achieve the performance gains of a fixed ANN topology (allowing for disciplined memory access) while searching a range of topologies.
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=\textwidth]{ann-chromosome}
+	\caption{ANN structure and related chromosome design.  Each element in the chromosome is a real-valued number.}
+	\label{fig:ann-chromosome}
+\end{figure}
+
+The genetic mutation and crossover operations which we have implemented are based on the methods defined by Montana and Lawrence \cite{GA-ANN}.
+Of the operations presented in this paper, the MUTATE NODES and CROSSOVER WEIGHTS operations were determined to provide an optimal balance between performance and classifier accuracy, and have been implemented in all of our experiments.
+
+% ----------------------------------------------------------------------------
+\subsection{Preliminary Data Analysis} \label{prelim}
+% ----------------------------------------------------------------------------
+The maximization of the ratio of between-class scatter to within-class scatter can be posed as a generalized eigenvalue problem where the eigenvalue with the largest corresponding eigenvalue is Fisher's linear discriminant \cite[\S 3.8.2]{Duda}.
+By taking instead the \emph{two} highest eigenvalues and plotting a 2D projection, we found that the data exhibited 4 defined clusters.
+This  guided our choice of 4 hidden nodes with Gaussian radial basis functions.
+
+% ----------------------------------------------------------------------------
+\subsection{Data Preprocessing Steps} \label{preprocessing}
+% ----------------------------------------------------------------------------
+In order to handle nominal feature values in data sets, we have implemented an algorithm which, given a tab seperated values file, converts nominal feature values in the data set into numeric values.
+The algorithm replaces each nominal value in the data set with the logarithm of the probability ratio between the positive and negative values observed given the nominal value.
+As a result, meaningful numeric values can be extracted from nominal values without domain knowledge relating to the values.
+This improves on the traditional orthogonal coding method \cite[slide 17]{MLP-NominalCoding} by avoiding allocation of a disproportionate number of input nodes to nominal features.
+
+% ############################################################################
 \section{Implementation of GA Training of ANNs} \label{algimp}
@@
 % ############################################################################
+We first present the computational performance of our algorithm for classifier training, followed by the estimated classification performance of the classifier designed for the UC San Diego data mining contest.
 
@@
 
+% ----------------------------------------------------------------------------
+\subsection{Classification Performance} \label{results}
+% ----------------------------------------------------------------------------
+The classifier was trained over 12000 generations; the fitness metric during this training was lift at 20\% calculated on a training set which consisted of 70\% of the data set, chosen at random during initialization.
+After each generation, the candidate ANN with the highest fitness value was used to classify 100 random subsets of the hold-out data set, and the average of the lift at 20\% of these classifications was recorded.
+Figure~\ref{fig:evolution-lift} demonstrates experimental results across 12,000 generations of population 50.
+The final estimated lift is 4.51 out of a maximum lift of 5.
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.5\textwidth]{fig-evolution-lift}
+	\caption{Evolution of classifier lift. Note logarithmic generation number axis.}
+	\label{fig:evolution-lift}
+\end{figure}
+
+When applied to the 2009 UC San Diego Data Mining Contest blind test data set, the resulting classifier produced a lift at 20\% value of 4.031, placing 35th in the world for classification of this data set.
+Due to the stochastic nature of genetic algorithms and the contest rules which state that contestants may test any number of candidate classifiers, it is likely that repeated entry into this contest would result in a higher ranking.
 % ############################################################################
@@
 Our method scales across population and training set sizes and is expected to be useful in other data intensive machine learning and data mining tasks.
+
+% ############################################################################
+\section{Future Work} \label{future}
+% ############################################################################
+Montana and Lawrence \cite{GA-ANN} have demonstrated that a brief period of hill climbing after training with a genetic algorithm may increase the accuracy of classifier results.
+Thus, we could improve the accuracy of our methods by incorporating hill-climbing methods, such as the parallel backpropagation method described by Oei, Friedland and Janin \cite{backprop}.
+
+In addition, parallelization methods of genetic algorithms involving multiple populations \cite{Alba1999} may interact favourably with data mining applications.
+Seperate populations may be trained on different subsets of the training data, allowing for a more thorough search of the solution space which the training set presents.
 

### 2010-04-21 doc/report.tex
Rewrite Preliminary Data Analysis and add scatter plot of our 2D projection
git-svn-id: http://svn.scesoc.ca/evolution/trunk@111 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 % ----------------------------------------------------------------------------
-The maximization of the ratio of between-class scatter to within-class scatter can be posed as a generalized eigenvalue problem where the eigenvalue with the largest corresponding eigenvalue is Fisher's linear discriminant \cite[\S 3.8.2]{Duda}.
-By taking instead the \emph{two} highest eigenvalues and plotting a 2D projection, we found that the data exhibited 4 defined clusters.
-This  guided our choice of 4 hidden nodes with Gaussian radial basis functions.
+To get a first idea of the separability of the data, we took an approach similar to Fisher's linear discriminant.
+This discriminant finds the 1D projection of the data which maximizes class separation, defined as the ratio of between-class variance and within-class variance.
+The problem of finding Fisher's linear discriminant can be seen as a generalized eigenvalue problem where the eigenvector with the large eigenvalue is the desired projection \cite[\S 3.8.2]{Duda}.
+To gain more insight into the higher-dimensional properties of the data, and because 2D visualization is practical, we instead used the eigenvectors with the \emph{two} highest eigenvalues and formed a 2D projection of the data (Figure~\ref{fig:fishers-2d}).
+The resulting four approximately-Gaussian clusters guided our decision to use a radial basis function network with four Gaussian hidden nodes.
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=\textwidth]{fishers_2d}
+	\caption{Projection of original data set on hyperplane formed by the two eigenvectors with the largest eigenvalues.}
+	\label{fig:fishers-2d}
+\end{figure}
 

### 2010-04-21 doc/report.tex
Add two paragraphs on the degenerate output values phenomenon and how we solved it (standardization of feature values)
git-svn-id: http://svn.scesoc.ca/evolution/trunk@112 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 
-The classifier receives as inputs one numeric value for each feature in a data set, all of which are normalized to a value range between 0 and 1.
-The classifier outputs a single numeric value, also in the range of 0 and 1, which represents the confidence with which the pattern is a member of the positive class.
+The classifier receives as inputs real-valued features which are standardized prior to processing by the ANN.
+The classifier outputs a single real value, in the range of 0 and 1, which represents the confidence with which the pattern is a member of the positive class.
 These confidence values can be used for ranking during lift calculations, or for sensitivity or specificity measures with an arbitrary value chosen as a threshold for positive identification of a pattern.
@@
 In our final design, each candidate ANN is represented by a chromosome which consists of four arrays of real-valued numbers.
-Edge weights between input nodes and hidden layers, edge weights between hidden and output layers, width values for RBFs at hidden nodes and center values for RBFs at hidden nodes are all encoded, and can take on any value.
+Edge weights between input nodes and hidden layers, edge weights between hidden and output layers, width values for RBFs at hidden nodes and center values for RBFs at hidden nodes are all encoded, and can take on any real value.
 Figure~\ref{fig:evolution-lift} demonstrates the chromosome structure and its relation to ANN structure.
@@
 This improves on the traditional orthogonal coding method \cite[slide 17]{MLP-NominalCoding} by avoiding allocation of a disproportionate number of input nodes to nominal features.
+
+Due to the inherent limitations of floating-point (FP) representation of real values, the range of inputs to the ANN plays a crucial part in the numerical stability of the classifier during training and during classification.
+In particular, our use of exponential functions resulted in an ANN very sensitive to output saturation.
+In these cases, while the input to the \texttt{expf} function is well within the range of representible FP numbers (take for example feature 11 which has a maximum of 8193), the mathematical output of the exponentiation is too large to be represented in FP.
+This causes degenerate output values such as infinity or NaN (not a number).
+These special values propagate through the ANN edges to produce degenerate values at the output of the classifier.
+
+To avoid this issue, we standardize the data before training and before classification.
+For each feature individually, we estimate the 10\textsuperscript{th} and 90\textsuperscript{th} percentiles of the feature values.
+Then we shift and scale the values such that 80\% of the resulting values are found between -1 and 1.
+This standardization method was chosen for its robustness against outliers.
+After standardization, the range of inputs to \texttt{expf} is narrow enough to always produce a valid FP value at the output.
+The standardization does not affect the classification itself because there are scaling and shifting parameters within the ANN (input to hidden edge weights and hidden center parameters, respectively) which are already part of the search space.
 

### 2010-04-21 doc/report.tex
More changes to the report
git-svn-id: http://svn.scesoc.ca/evolution/trunk@113 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 The classifier was trained over 12000 generations; the fitness metric during this training was lift at 20\% calculated on a training set which consisted of 70\% of the data set, chosen at random during initialization.
-After each generation, the candidate ANN with the highest fitness value was used to classify 100 random subsets of the hold-out data set, and the average of the lift at 20\% of these classifications was recorded.
+After each generation, the candidate ANN with the highest fitness value was used to classify 100 random subsets of the hold-out data set (each consisting of exactly one half of the hold-out data set), and the average of the lift at 20\% of these classifications was recorded.
 Figure~\ref{fig:evolution-lift} demonstrates experimental results across 12,000 generations of population 50.
@@
 Due to the stochastic nature of genetic algorithms and the contest rules which state that contestants may test any number of candidate classifiers, it is likely that repeated entry into this contest would result in a higher ranking.
+The discrepency between the estimated performance and actual performance on the blind test data set can be attributed to the following:
+\begin{enumerate}
+	\item Discrepency in test set size: the estimated performance measurements were done on sets of approximately 14000 patterns while the blind test data set has approximately 36000 patterns.
+	\item Discrepency in class balance: when splitting the data set into training and hold-out data sets, class balance was not enforced for either of the sets.
+The blind test data set was known to have the same class imbalance as the given data set.
+Therefore, it is possible that the data used to collect experimental data had a class imbalance which did not reflect the blind test data set.
+	\item Nominal values unique to the blind test data set: It is possible that patterns with nominal values that do not appear in the training data set could cause misclassification, however, the pre-processing step which converts nominal values into numeric values accounts for values which have not been seen previously.
+\end{enumerate}
+
 % ############################################################################

### 2010-04-21 doc/report.tex
SIGNING OFF HOPEFULLY NOT FOR THE WHOLE NIGHT, GOOD LUCK
git-svn-id: http://svn.scesoc.ca/evolution/trunk@114 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 GPUs, however, use a specialized programming paradigm which must be taken into account to leverage their full processing power.
-In this paper, we show an implementation of GA training of ANNs which achieves an order-of-magnitude speedup over a sequential algorithm.
+In this report, we show an implementation of GA training of ANNs which achieves an order-of-magnitude speedup over a sequential algorithm.
 
@@
 Each layer is fully connected to the next.
+Figure~\ref{fig:ann-chromosome} demonstrates the feedfoward ANN structure.
 Each of the hidden nodes at hidden layer $n$ perform a transcendental function with inputs equal to the outputs of layer $n-1$ each multiplied by the weight associated with the edge through which they are fed from layer $n-1$ to layer $n$.
@@
 % ############################################################################
-\subsection{Classifier Design} \label{design}
+\section{Classifier Design} \label{design}
 % ############################################################################
 
-While our GA-ANN classifier was designed for the 2009 UC San Diego Data Mining Contest data set, it is capable of adapting to a variety of data sets.
-We have developped a flexible ANN topology, allowing for one or two hidden layers and an arbitrary number of nodes per hidden layer (within the limits of GPU memory).
+% ----------------------------------------------------------------------------
+\subsection{ANN Structure Design} \label{ann}
+% ----------------------------------------------------------------------------
+
+While our GA-ANN classifier was designed for the 2009 UC San Diego Data Mining Contest data set, it has been designed in order to be capable of adapting to a variety of data sets.
+We have developed a flexible ANN topology, allowing for either one or two layers of hidden nodes and an arbitrary number of hidden nodes per layer (within the limits of GPU memory).
 In addition, each layer of the ANN is capable of performing RBF or sigmoid calculations.
@@
 These confidence values can be used for ranking during lift calculations, or for sensitivity or specificity measures with an arbitrary value chosen as a threshold for positive identification of a pattern.
-Therefore, the fitness function is not limited to lift at 20\% as presented in this paper.
+Therefore, the fitness function is not limited to lift at 20\% as presented in this report.
 
-In our final design, each candidate ANN is represented by a chromosome which consists of four arrays of real-valued numbers.
+% ----------------------------------------------------------------------------
+\subsection{Genetic Algorithm Design} \label{ann}
+% ----------------------------------------------------------------------------
+
+For a single-layer ANN, as used in our final design, each candidate ANN is represented by a chromosome which consists of four arrays of real-valued numbers.
 Edge weights between input nodes and hidden layers, edge weights between hidden and output layers, width values for RBFs at hidden nodes and center values for RBFs at hidden nodes are all encoded, and can take on any real value.
-Figure~\ref{fig:evolution-lift} demonstrates the chromosome structure and its relation to ANN structure.
+Figure~\ref{fig:ann-chromosome} demonstrates the chromosome structure and its relation to ANN structure.
 This structure is simple yet powerful; individual edges are 'removed' by the genetic algorithm when their weight is set to zero and nodes are 'removed' by the genetic algorithm when their RBF width value is set to zero.
-Thus, we achieve the performance gains of a fixed ANN topology (allowing for disciplined memory access) while searching a range of topologies.
+Thus, while the topology of the ANN is limited to those with $m$ layers and $n$ nodes per layer, the genetic algorithm searches all topologies with $m$ layers and between 0 and $n$ nodes per layer.
 
@@
 In order to handle nominal feature values in data sets, we have implemented an algorithm which, given a tab seperated values file, converts nominal feature values in the data set into numeric values.
-The algorithm replaces each nominal value in the data set with the logarithm of the probability ratio between the positive and negative values observed given the nominal value.
+The algorithm replaces each nominal value in the data set with the logarithm of the probability ratio between the positive and negative classifications observed given the nominal value.
 As a result, meaningful numeric values can be extracted from nominal values without domain knowledge relating to the values.
 This improves on the traditional orthogonal coding method \cite[slide 17]{MLP-NominalCoding} by avoiding allocation of a disproportionate number of input nodes to nominal features.
+The logarithm of the ratio of all positive classifications to all negative classifications is attributed to nominal values which do not appear sufficiently in the data set and for values which appear only in blind test data sets; this prevents the classifier from attributing significant values to nominal values for which little or no inference can be done.
 
@@
 The discrepency between the estimated performance and actual performance on the blind test data set can be attributed to the following:
-\begin{enumerate}
+\begin{itemize}
 	\item Discrepency in test set size: the estimated performance measurements were done on sets of approximately 14000 patterns while the blind test data set has approximately 36000 patterns.
 	\item Discrepency in class balance: when splitting the data set into training and hold-out data sets, class balance was not enforced for either of the sets.
-The blind test data set was known to have the same class imbalance as the given data set.
-Therefore, it is possible that the data used to collect experimental data had a class imbalance which did not reflect the blind test data set.
+The blind test data set was known to have the same class balance as the given data set.
+Therefore, it is possible that the hold-out data set used to collect experimental results had a class balance which did not reflect the blind test data set.
 	\item Nominal values unique to the blind test data set: It is possible that patterns with nominal values that do not appear in the training data set could cause misclassification, however, the pre-processing step which converts nominal values into numeric values accounts for values which have not been seen previously.
-\end{enumerate}
+\end{itemize}
 

### 2010-04-22 doc/report.tex
Small tweaks in wording
git-svn-id: http://svn.scesoc.ca/evolution/trunk@115 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 This discriminant finds the 1D projection of the data which maximizes class separation, defined as the ratio of between-class variance and within-class variance.
-The problem of finding Fisher's linear discriminant can be seen as a generalized eigenvalue problem where the eigenvector with the large eigenvalue is the desired projection \cite[\S 3.8.2]{Duda}.
+The problem of finding Fisher's linear discriminant can be seen as a generalized eigenvalue problem where the eigenvector with the largest corresponding eigenvalue is the desired projection \cite[\S 3.8.2]{Duda}.
 To gain more insight into the higher-dimensional properties of the data, and because 2D visualization is practical, we instead used the eigenvectors with the \emph{two} highest eigenvalues and formed a 2D projection of the data (Figure~\ref{fig:fishers-2d}).
@@
 % ----------------------------------------------------------------------------
-In order to handle nominal feature values in data sets, we have implemented an algorithm which, given a tab seperated values file, converts nominal feature values in the data set into numeric values.
+In order to handle nominal feature values in data sets, we have implemented an algorithm which, given a tab separated values (TSV) file, converts nominal feature values in the data set into numeric values.
 The algorithm replaces each nominal value in the data set with the logarithm of the probability ratio between the positive and negative classifications observed given the nominal value.
+This is inspired from the discriminant function of a Bayesian classifier trained only on that particular feature.
 As a result, meaningful numeric values can be extracted from nominal values without domain knowledge relating to the values.
 This improves on the traditional orthogonal coding method \cite[slide 17]{MLP-NominalCoding} by avoiding allocation of a disproportionate number of input nodes to nominal features.
-The logarithm of the ratio of all positive classifications to all negative classifications is attributed to nominal values which do not appear sufficiently in the data set and for values which appear only in blind test data sets; this prevents the classifier from attributing significant values to nominal values for which little or no inference can be done.
+The logarithm of the ratio of all positive classifications to all negative classifications is assigned to nominal values which do not appear sufficiently in the data set and for values which appear only in blind test data sets; this prevents the classifier from assigning significant values to nominal values for which little or no inference can be done.
 
@@
 For each feature individually, we estimate the 10\textsuperscript{th} and 90\textsuperscript{th} percentiles of the feature values.
-Then we shift and scale the values such that 80\% of the resulting values are found between -1 and 1.
+We then shift and scale the values such that 80\% of the resulting values are in the range $(-1, 1)$.
 This standardization method was chosen for its robustness against outliers.
@@
 Feature values for all training instances are loaded into GPU global memory in preparation for ANN computations.
-Then, the following steps are executed in sequence for each generation of the GA:
+The following steps are then executed in sequence for each generation of the GA:
 
@@
 The classifier was trained over 12000 generations; the fitness metric during this training was lift at 20\% calculated on a training set which consisted of 70\% of the data set, chosen at random during initialization.
-After each generation, the candidate ANN with the highest fitness value was used to classify 100 random subsets of the hold-out data set (each consisting of exactly one half of the hold-out data set), and the average of the lift at 20\% of these classifications was recorded.
+After each generation, the candidate ANN with the highest fitness value was used to classify 20 random subsets of the hold-out data set (each consisting of exactly one half of the hold-out data set), and the average of the lift at 20\% of these classifications was recorded.
 Figure~\ref{fig:evolution-lift} demonstrates experimental results across 12,000 generations of population 50.
@@
 
-When applied to the 2009 UC San Diego Data Mining Contest blind test data set, the resulting classifier produced a lift at 20\% value of 4.031, placing 35th in the world for classification of this data set.
+When applied to the 2009 UC San Diego Data Mining Contest blind test data set, the resulting classifier produced a lift at 20\% value of 4.031, placing 35\textsuperscript{th} in the world for classification of this data set.
 Due to the stochastic nature of genetic algorithms and the contest rules which state that contestants may test any number of candidate classifiers, it is likely that repeated entry into this contest would result in a higher ranking.
-The discrepency between the estimated performance and actual performance on the blind test data set can be attributed to the following:
+The discrepancy between the estimated performance and actual performance on the blind test data set can be attributed to the following:
 \begin{itemize}
-	\item Discrepency in test set size: the estimated performance measurements were done on sets of approximately 14000 patterns while the blind test data set has approximately 36000 patterns.
-	\item Discrepency in class balance: when splitting the data set into training and hold-out data sets, class balance was not enforced for either of the sets.
+	\item Discrepancy in test set size: the estimated performance measurements were done on sets of approximately 14000 patterns while the blind test data set has approximately 36000 patterns.
+	\item Discrepancy in class balance: when splitting the data set into training and hold-out data sets, class balance was not enforced for either of the sets.
 The blind test data set was known to have the same class balance as the given data set.
 Therefore, it is possible that the hold-out data set used to collect experimental results had a class balance which did not reflect the blind test data set.
-	\item Nominal values unique to the blind test data set: It is possible that patterns with nominal values that do not appear in the training data set could cause misclassification, however, the pre-processing step which converts nominal values into numeric values accounts for values which have not been seen previously.
+	\item Nominal values unique to the blind test data set: It is possible that patterns with nominal values that do not appear in the training data set could cause misclassification.
+However, the pre-processing step which converts nominal values into numeric values accounts for values which are not previously seen.
 \end{itemize}

### 2010-06-29 bench_plot.m
Check in perf plotting script which takes "generic x86" data from Core2 (more data points than i7)
git-svn-id: http://svn.scesoc.ca/evolution/trunk@118 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@ -1,0 +1,31 @@
+fp = fopen('bench_wrt_popsize.txt');
+data = textscan(fp, '%12n%12n%12n%12n%12n%12n', ...
+  'HeaderLines', 1, 'ReturnOnError', 0);
+fclose(fp);
+data = cell2mat(data);
+
+subplot(1,2,1);
+loglog(data(:,1), data(:,3) * 1e-3, '^-', ...
+       data(:,1), data(:,5) * 1e-3, '^--', ...
+       data(:,1), data(:,6) * 1e-3, 's--');
+title({'One Generation with', '10000 Training Instances'}, ...
+  'FontSize',12,'FontWeight','b');
+legend('x86', 'GTX260', 'GTX275', 'Location', 'NorthWest');
+xlabel('Population size (number of ANNs)');
+ylabel('Execution time (s)');
+
+fp = fopen('bench_wrt_trainsize.txt');
+data = textscan(fp, '%12n%12n%12n%12n%12n%12n', ...
+  'HeaderLines', 1, 'ReturnOnError', 0);
+fclose(fp);
+data = cell2mat(data);
+
+subplot(1,2,2);
+loglog(data(:,2), data(:,3) * 1e-3, '^-', ...
+       data(:,2), data(:,5) * 1e-3, '^--', ...
+       data(:,2), data(:,6) * 1e-3, 's--');
+title({'One Generation of', '100 ANNs'}, ...
+  'FontSize', 12, 'FontWeight', 'b');
+legend('x86', 'GTX260', 'GTX275', 'Location', 'NorthWest');
+xlabel('Training set size (number of instances)');
+%ylabel('Execution time (s)');

### 2010-06-29 doc/HPCS-proceedings-paper.tex
git-svn-id: http://svn.scesoc.ca/evolution/trunk@120 fee98b4c-d1c3-4d34-9525-a468159e4a48

--- 2010-04-22 doc/report.tex
@@
 \evensidemargin=0.25in \topmargin=-0.5in
-\linespread{2}
 %--------------- Algorithm --------------------------------------------------
@@
 % ----------------------------------------------------------------------------
-\subsection{Preliminary Data Analysis} \label{prelim}
-% ----------------------------------------------------------------------------
-To get a first idea of the separability of the data, we took an approach similar to Fisher's linear discriminant.
-This discriminant finds the 1D projection of the data which maximizes class separation, defined as the ratio of between-class variance and within-class variance.
-The problem of finding Fisher's linear discriminant can be seen as a generalized eigenvalue problem where the eigenvector with the largest corresponding eigenvalue is the desired projection \cite[\S 3.8.2]{Duda}.
-To gain more insight into the higher-dimensional properties of the data, and because 2D visualization is practical, we instead used the eigenvectors with the \emph{two} highest eigenvalues and formed a 2D projection of the data (Figure~\ref{fig:fishers-2d}).
-The resulting four approximately-Gaussian clusters guided our decision to use a radial basis function network with four Gaussian hidden nodes.
-
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=\textwidth]{fishers_2d}
-	\caption{Projection of original data set on hyperplane formed by the two eigenvectors with the largest eigenvalues.}
-	\label{fig:fishers-2d}
-\end{figure}
-
-% ----------------------------------------------------------------------------
 \subsection{Data Preprocessing Steps} \label{preprocessing}
@@
 	\hline
-	Core2  & Intel Core 2 Q9450 CPU & 2.66~GHz & 7.5~GB & Linux 2.6 (64-bit) \\
-	i7     & Intel i7 920 CPU       & 2.66~GHz & 2.4~GB & Windows 7 (64-bit) \\
+	x86    & x86 Desktop Processor  & 2.66~GHz & 7.5~GB & Linux 2.6 (64-bit) \\
 	GTX260 & NVidia GTX260 GPU      & 1.24~GHz & N/A    & Windows 7 (64-bit) \\
@@
 % ----------------------------------------------------------------------------
-In most cases, the GPU implementation resulted in approximately an order of magnitude speedup over a CPU implementation (Figure~\ref{fig:training-performance}).
+In most cases, the GPU implementation resulted in approximately an order of magnitude speedup over a serial implementation of equivalent code running on a modern x86 desktop processor (Figure~\ref{fig:training-performance}).
 For population sizes below 10, observed speedup decreases and no speedup is observed in the single-candidate case.
@@
 	\centering
-	\includegraphics[width=\textwidth]{fig-performance}
+	\includegraphics[width=\textwidth]{fig-performance-genericx86}
 	\caption{Training performance as a function of population size (left) and training set size (right). Note logarithmic axes.}
@@
 \end{figure}
-
-When applied to the 2009 UC San Diego Data Mining Contest blind test data set, the resulting classifier produced a lift at 20\% value of 4.031, placing 35\textsuperscript{th} in the world for classification of this data set.
-Due to the stochastic nature of genetic algorithms and the contest rules which state that contestants may test any number of candidate classifiers, it is likely that repeated entry into this contest would result in a higher ranking.
-The discrepancy between the estimated performance and actual performance on the blind test data set can be attributed to the following:
-\begin{itemize}
-	\item Discrepancy in test set size: the estimated performance measurements were done on sets of approximately 14000 patterns while the blind test data set has approximately 36000 patterns.
-	\item Discrepancy in class balance: when splitting the data set into training and hold-out data sets, class balance was not enforced for either of the sets.
-The blind test data set was known to have the same class balance as the given data set.
-Therefore, it is possible that the hold-out data set used to collect experimental results had a class balance which did not reflect the blind test data set.
-	\item Nominal values unique to the blind test data set: It is possible that patterns with nominal values that do not appear in the training data set could cause misclassification.
-However, the pre-processing step which converts nominal values into numeric values accounts for values which are not previously seen.
-\end{itemize}
 

### 2010-06-30 doc/HPCS-proceedings-paper.tex
New version of the paper
git-svn-id: http://svn.scesoc.ca/evolution/trunk@122 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 Ottawa, Canada K1S 5B6\\
-{\em cat@vv.carleton.ca}, {\em rpeace@sce.carleton.ca}, {\em jrgreen@sce.carleton.ca}
+{\em cpatulea@sce.carleton.ca}, {\em rpeace@sce.carleton.ca}, {\em jrgreen@sce.carleton.ca}
 % ############################################################################
@@
 However, because they require the evaluation of approximately 50 to 100 classifiers at each iteration, and the algorithm is repeated for several thousand iterations, they are extremely compute-intensive.
-This is exacerbated by the use of artificial neural networks (ANNs) as the classifier because ANNs themselves are very compute-intensive.
+The use of artificial neural networks (ANNs) as the classifier exacerbates the computational requirements because ANNs themselves are very compute-intensive.
 
@@
 ANN training has been implemented previously on GPU hardware, however these efforts have focused on training through backpropagation \cite{backprop} as opposed to genetic algorithms.
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=\textwidth]{ann-chromosome}
+	\caption{ANN structure and related chromosome design.  Each element in the chromosome is a real-valued number.}
+	\label{fig:ann-chromosome}
+\end{figure}
 
@@
 
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=\textwidth]{ann-chromosome}
-	\caption{ANN structure and related chromosome design.  Each element in the chromosome is a real-valued number.}
-	\label{fig:ann-chromosome}
-\end{figure}
-
 The genetic mutation and crossover operations which we have implemented are based on the methods defined by Montana and Lawrence \cite{GA-ANN}.
@@
 In particular, our use of exponential functions resulted in an ANN very sensitive to output saturation.
-In these cases, while the input to the \texttt{expf} function is well within the range of representible FP numbers (take for example feature 11 which has a maximum of 8193), the mathematical output of the exponentiation is too large to be represented in FP.
+In these cases, while the input to the \texttt{expf} function is well within the range of representible FP numbersthe mathematical output of the exponentiation is too large to be represented in FP.
 This causes degenerate output values such as infinity or NaN (not a number).
@@
 For population sizes below 10, observed speedup decreases and no speedup is observed in the single-candidate case.
-This is due to GPU initialization and data copying dominating execution time.
+This is due to GPU kernel invocation overhead and data copying dominating execution time.
 However, in these cases, both implementations are still quite fast (20~ms per generation).

### 2010-07-09 doc/HPCS-proceedings-paper.tex
Formatted report to meet JPCS requirements
git-svn-id: http://svn.scesoc.ca/evolution/trunk@123 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 % ===========================================================================
-\documentclass[11pt]{article}       %--- LATEX 2e base
+\documentclass[letterpaper]{jpconf}       %--- LATEX 2e base
 \usepackage{latexsym}               %--- LATEX 2e base
@@
 
-\title{CUDA-accelerated Genetic Feedforward-ANN Training for Data Mining}
+\title{CUDA-accelerated genetic feedforward-ANN training for data mining}
 
@@
 % Author(s) (no blank lines !)
-\author{
+\author{Catalin Patulea, Robert Peace and James Green}
+\address{School of Systems and Computer Engineering,
+Carleton University,
+Ottawa, Canada K1S 5B6}
+\ead{cpatulea@sce.carleton.ca, rpeace@sce.carleton.ca, jrgreen@sce.carleton.ca}
 % ############################################################################
-Catalin Patulea, Robert Peace, and James Green\\
-School of Systems and Computer Engineering\\
-Carleton University\\
-Ottawa, Canada K1S 5B6\\
-{\em cpatulea@sce.carleton.ca}, {\em rpeace@sce.carleton.ca}, {\em jrgreen@sce.carleton.ca}
+% end-authors
 % ############################################################################
-} % end-authors
-% ############################################################################
-
-\maketitle
 
@@
 In addition, parallelization methods of genetic algorithms involving multiple populations \cite{Alba1999} may interact favourably with data mining applications.
-Seperate populations may be trained on different subsets of the training data, allowing for a more thorough search of the solution space which the training set presents.
+Seperate populations may be trained on different subsets of the training data, allowing for a more thorough search of the solution space which the training set presents.\\
 
@@
 % ############################################################################
-\bibliographystyle{unsrt}
+\bibliographystyle{iopart-num}
 \bibliography{bibliography}     %loads bibliography.bib

### 2010-07-09 doc/HPCS-proceedings-paper.tex
Formatted the tables and figures
git-svn-id: http://svn.scesoc.ca/evolution/trunk@124 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 \begin{figure}[h]
-	\centering
+	\begin{center}
 	\includegraphics[width=\textwidth]{ann-chromosome}
-	\caption{ANN structure and related chromosome design.  Each element in the chromosome is a real-valued number.}
-	\label{fig:ann-chromosome}
+	\caption{\label{fig:ann-chromosome}ANN structure and related chromosome design.  Each element in the chromosome is a real-valued number.}
+	\end{center}
 \end{figure}
@@
 \begin{table}[h]
-	\centering
+	\caption{\label{tab:experimental-hardware}Experimental hardware. Free RAM is OS-reported available memory before running each experiment.}
+	\begin{center}
 	\begin{tabular}{lllll}
-	\textbf{Legend Label} & \textbf{Hardware} & \textbf{Clock} & \textbf{Free RAM} & \textbf{OS} \\
-	\hline
-	x86    & x86 Desktop Processor  & 2.66~GHz & 7.5~GB & Linux 2.6 (64-bit) \\
-	GTX260 & NVidia GTX260 GPU      & 1.24~GHz & N/A    & Windows 7 (64-bit) \\
-	GTX275 & NVidia GTX275 GPU      & 1.4~GHz  & N/A    & Linux 2.6 (64-bit) \\
-	\hline
+	\br
+	Legend Label&Hardware&Clock&Free RAM&OS\\
+	\mr
+	x86   &x86 Desktop Processor&2.66~GHz&7.5~GB&Linux 2.6 (64-bit) \\
+	GTX260&NVidia GTX260 GPU    &1.24~GHz&N/A   &Windows 7 (64-bit) \\
+	GTX275&NVidia GTX275 GPU    &1.4~GHz &N/A   &Linux 2.6 (64-bit) \\
+	\br
 	\end{tabular}
-	\caption{Experimental hardware. Free RAM is OS-reported available memory before running each experiment.}
-	\label{tab:experimental-hardware}
+	\end{center}
 \end{table}
@@
 \begin{figure}[h]
-	\centering
+	\begin{center}
 	\includegraphics[width=\textwidth]{fig-performance-genericx86}
-	\caption{Training performance as a function of population size (left) and training set size (right). Note logarithmic axes.}
-	\label{fig:training-performance}
+	\caption{\label{fig:training-performance}Training performance as a function of population size (left) and training set size (right). Note logarithmic axes.}
+	\end{center}
 \end{figure}
@@
 \begin{figure}[h]
-	\centering
+	\begin{center}
 	\includegraphics[width=0.5\textwidth]{fig-evolution-lift}
-	\caption{Evolution of classifier lift. Note logarithmic generation number axis.}
-	\label{fig:evolution-lift}
+	\caption{\label{fig:evolution-lift}Evolution of classifier lift. Note logarithmic generation number axis.}
+	\end{center}
 \end{figure}

### 2010-07-09 doc/HPCS-proceedings-paper.tex
Add references and further explanation on method of standardization of feature values
git-svn-id: http://svn.scesoc.ca/evolution/trunk@125 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 
-To avoid this issue, we standardize the data before training and before classification.
-For each feature individually, we estimate the 10\textsuperscript{th} and 90\textsuperscript{th} percentiles of the feature values.
-We then shift and scale the values such that 80\% of the resulting values are in the range $(-1, 1)$.
-This standardization method was chosen for its robustness against outliers.
-After standardization, the range of inputs to \texttt{expf} is narrow enough to always produce a valid FP value at the output.
+To avoid this issue, we standardize each feature individually before training and before classification.
+One method of standardization is to use the sample mean and sample standard deviation as shifting and scaling parameters (referred to as "z-score normalization" in \cite{ZscoreNormalization}).
+However, sample standard deviation is not robust against outliers, while interquartile range (IQR) does tolerate outliers \cite{RobustRegression}.
+We use a variant of IQR: we estimate the range between the 10\textsuperscript{th} and 90\textsuperscript{th} percentiles and use that as the scaling parameter.
+After shifting and scaling, 80\% of the resulting feature values are in the range $(-1, 1)$.
+This range of inputs to \texttt{expf} is narrow enough to always produce a valid FP value at the output.
 The standardization does not affect the classification itself because there are scaling and shifting parameters within the ANN (input to hidden edge weights and hidden center parameters, respectively) which are already part of the search space.

### 2010-07-09 doc/HPCS-proceedings-paper.tex
Small typo/wording fixes
git-svn-id: http://svn.scesoc.ca/evolution/trunk@126 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 In particular, our use of exponential functions resulted in an ANN very sensitive to output saturation.
-In these cases, while the input to the \texttt{expf} function is well within the range of representible FP numbersthe mathematical output of the exponentiation is too large to be represented in FP.
+In these cases, while the input to the \texttt{expf} function is well within the range of representible FP numbers the mathematical output of the exponentiation is too large to be represented in FP.
 This causes degenerate output values such as infinity or NaN (not a number).
@@
 
-Our GA begins by generating an initial population of random candidate ANNs, each of which is a set of parameters describing one ANN.
+Our GA begins by generating an initial population of random candidate ANNs, where each chromosome in the population is a set of parameters describing one ANN.
 Feature values for all training instances are loaded into GPU global memory in preparation for ANN computations.
@@
 	\item Compute the ANN output corresponding to each training instance, as calculated by each candidate ANN in the population.
-	\item Find the threshold that defines the highest 20\% outputs for each candidate's set of output values.
+	\item Find the threshold that defines the outputs which fall in the top 20\% for each candidate ANN's set of output values.
 	\item Compute the number of top 20\% instances which are truly positive, again for each candidate ANN. This allows us to calculate lift, which is used as the fitness value for each candidate ANN.

### 2010-07-10 doc/HPCS-proceedings-paper.tex
Add clarification to computation of ANN values
git-svn-id: http://svn.scesoc.ca/evolution/trunk@127 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 % ----------------------------------------------------------------------------
-ANN output values are computed by reading the training instances as real-valued feature vectors and applying the appropriate operations represented by the network topology and parameters.
+ANN output values are computed by reading the training instances as real-valued feature vectors, applying them to the input nodes of the network (see Figure~\ref{fig:ann-chromosome}), and storing the output of the network.
+Specifically, each feature value is first scaled by the corresponding input-hidden edges.
+These values are then propagated to all hidden nodes, which apply a radial basis function.
+Finally, the hidden node outputs are scaled by the hidden-output edges and summed at the output node.
+A single real value is output from the network and stored in GPU memory.
+Parameters for edge weights and radial basis functions are given in the chromosome of each candidate ANN.
+
 Feature data are organized to efficiently use memory bandwidth.
+The requirement for optimal use of CUDA memory bandwidth is that blocks of threads simultaneously access sequential addresses in memory, resulting in a coalesced single-clock cycle access.
+We store feature data in feature-major order (values for all instances of feature 1, all instances of feature 2, etc.).
+Therefore, when the thread block reads data for the first feature across different instances, the memory accesses can be coalesced.
+Similarly, accesses are coalesced for all subsequent features.
+
 ANN output calculation is particularly well-suited to the CUDA architecture because it has a very high ratio of mathematical operations to memory accesses (160 multiplications, 4 exponentiations and 80 bytes of memory accesses for our ANN topology).

### 2010-07-10 doc/HPCS-proceedings-paper.tex
Specify the number of passes and adjust the percentiles
git-svn-id: http://svn.scesoc.ca/evolution/trunk@128 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 
-Because fast on-chip memory (16~KB) is not large enough for our value of $k$ ($4 * 18936 =$ 74~KB), we instead use $p$ passes of $\frac{k}{p}$-sized heaps.
+Because fast on-chip memory (16~KB) is not large enough for our value of $k$ ($4 * 18936 =$ 74~KB), we instead use $p$ (5) passes of $\frac{k}{p}$-sized heaps.
 Each pass examines only values which are less than the threshold calculated by the preceeding pass.
-For example, the third pass examines only values below the 10\% threshold and finds the top 5\% threshold of this subset of values.
-This threshold therefore is the 15\% threshold of the full list.
+For example, the third pass examines only values below the 8\% threshold and finds the top 4\% threshold of this subset of values.
+This threshold therefore is the 12\% threshold of the full list.
 The threshold calculated by the last pass is the desired 20\% threshold.

### 2010-07-10 doc/HPCS-proceedings-paper.tex
Rephrase in conclusion why multipass was required
git-svn-id: http://svn.scesoc.ca/evolution/trunk@129 fee98b4c-d1c3-4d34-9525-a468159e4a48
Reword how the top 20% counting is done
git-svn-id: http://svn.scesoc.ca/evolution/trunk@130 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 % ----------------------------------------------------------------------------
-To compute the number of truly positive instances in the top 20\% of scores, if training instances are unsorted, we must read from memory each instance's true class and corresponding ANN output value and compare the output value to the candidate's threshold.
+We wish to compute the number of truly positive instances in the top 20\% of scores.
+Assuming the training instances are unsorted, we must read from memory each instance's true class and corresponding ANN output value and compare the output value to the candidate's threshold.
 For $n$ training instances, this requires $2 * n$ memory accesses.
-However, by keeping the $n_p$ positive instances first in memory, we need only read the first $n_p$ ANN outputs, knowing that they correspond to the positive instances.
+However, by pre-sorting the training data such that the $n_p$ positive instances appear first in memory, we need only read the first $n_p$ ANN outputs, knowing that they correspond to the positive instances.
 This reduces the number of memory accesses to $n_p$, which is significant particularly for training data with low positive rate (in our case, $2 * n = 189364$ while $n_p = 2080$).
@@
 
-Counting of positive instances above the candidate's threshold is performed independently, in parallel, for each candidate.
+Counting of positive instances above the candidate's threshold is performed independently, in parallel, for each candidate ANN.
 However, because there may not be enough candidates to fully occupy the GPU, we also split the counting process for each candidate into 256 independent sub-counts: the first sub-counter processes training instances 0, 256, 512, \ldots, the second sub-counter instances 1, 257, 513, \ldots, such that the sub-counts can be executed in parallel.
@@
 By carefully designing memory organization, algorithm computational load and memory access patterns, we have obtained a 10-fold speedup compared to a conventional sequential CPU implementation.
-A multipass approach was required for part of our system to compensate for the limited amount of fast on-chip GPU memory.
+A multipass approach was required in the selection step of classifier evaluation to compensate for the limited amount of fast on-chip GPU memory.
 Our method scales across population and training set sizes and is expected to be useful in other data intensive machine learning and data mining tasks.

### 2010-07-10 doc/HPCS-proceedings-paper.tex
Fixup blurb about the experimental hardware
git-svn-id: http://svn.scesoc.ca/evolution/trunk@131 fee98b4c-d1c3-4d34-9525-a468159e4a48

@@
 % ----------------------------------------------------------------------------
-To evaluate the performance of our system, we compared execution time of equivalent algorithms running on commodity desktop CPUs and GPUs.
+To evaluate the performance of our system, we compared execution time of a sequential algorithm on a commodity CPU and an optimized algorithm on commodity GPUs.
+The goal was to estimate the speedup of our algorithm with respect to sequential without using the CPU's multiprocessing capabilities.
 Each series represents one type of experimental hardware (Table~\ref{tab:experimental-hardware}).
@@
 	\br
-	Legend Label&Hardware&Clock&Free RAM&OS\\
+	Legend Label & Hardware & Clock & Free RAM & OS\\
 	\mr
-	x86   &x86 Desktop Processor&2.66~GHz&7.5~GB&Linux 2.6 (64-bit) \\
-	GTX260&NVidia GTX260 GPU    &1.24~GHz&N/A   &Windows 7 (64-bit) \\
-	GTX275&NVidia GTX275 GPU    &1.4~GHz &N/A   &Linux 2.6 (64-bit) \\
+	x86    & Intel Core 2 Q9450 CPU & 2.66~GHz & 7.5~GB & Linux 2.6 (64-bit) \\
+	GTX260 & NVidia GTX260 GPU      & 1.24~GHz & N/A    & Windows 7 (64-bit) \\
+	GTX275 & NVidia GTX275 GPU      & 1.4~GHz  & N/A    & Linux 2.6 (64-bit) \\
 	\br
